---
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    include-in-header:
      text: |
        <style>
        .reveal {
          font-size: 35px;
        }

        .title-slide {
          text-align: center;
          padding: 45px;
        }

        .main-title {
          font-size: 1.8em;
          font-weight: 700;
          color: #9f1f55;
          line-height: 1.15;
          margin-bottom: 30px;
        }

        .subtitle {
          font-size: 0.95em;
          color: #333333;
          margin-bottom: 45px;
        }

        .authors {
          font-size: 0.8em;
          color: #222222;
          line-height: 1.5;
        }

        .reveal h2 {
          color: #9f1f55;
        }

        .reveal .mermaid {
          font-size: 26px;
        }
        .model-flow {
          width: 80%;
          margin: 0 auto;
          text-align: center;
          font-weight: 700;
          line-height: 1.4;
        }
        .map-slide .html-widget {
          width: 100% !important;
          height: 520px !important;
        }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false


library(dplyr)
library(ggplot2)
library(readr)
library(readxl)
library(sf)
library(plotly)
library(leaflet)
library(janitor)
library(tidyr)
library(stringr)
```

## {.title-slide}

::: {.main-title}
Distance, Mobility, and Mode Choice
:::

::: {.subtitle}
A Spatial Analysis of School Travel Behaviour in Ireland
:::

::: {.authors}
**Judith Beltrán López**  
**Sananthaa Jagadesan Senthamaraikannan**
:::

## Background

School travel is not only a question of **individual choice** — it is also shaped by **distance, school location, and educational provision**.

- Walking and cycling to school can improve health, learning, traffic congestion, and emissions.
- In Ireland, low active travel is often explained as a need for “behaviour change”.
- However, active travel does **not simply follow population density**.
- Instead, it is linked to the spatial distribution of schools and inequalities in access.


## Objectives

::: {.incremental}
1. Map and analyse the spatial distribution of schools across Ireland.

2. Examine the distance between students’ residential locations and their schools, and investigate student mobility related to school attendance.

3. Analyse school travel behaviour and assess the prevalence of healthy versus less healthy commuting modes.

4. Explore the relationship between school proximity, travel mode, and potential health implications.
:::

## Datasets: School Data

The school dataset comes from the **Irish Department of Education** and provides information for individual schools.

::: {.incremental}
- Includes data for **primary** and **post-primary** schools.
- Contains information on **enrolment figures** by school.
- Primary school data also includes **class size information**.
- Post-primary data includes specific categories such as **fee-paying** and **boarding schools**.
- This dataset helps describe the **spatial distribution and size of schools** across Ireland.
:::

## Datasets: School Data {.map-slide}

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyr)
library(leaflet)
schools_df <- readr::read_csv("data/School_Data_Eircode_2024_2025.csv",
  locale = readr::locale(encoding = "Latin1"),
  show_col_types = FALSE
) |>
  janitor::clean_names()

schools_sf <- schools_df |>
  dplyr::filter(!is.na(school_latitude), !is.na(school_longitude)) |>
  sf::st_as_sf(coords = c("school_longitude", "school_latitude"), crs = 4326, remove = FALSE)

enrol <- schools_sf$total_2024_2025
enrol[is.na(enrol)] <- 0
radius <- scales::rescale(enrol, to = c(2, 9))

ireland_counties <- sf::st_read("data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE) |>
  sf::st_transform(4326)

county_counts <- schools_df |>
  janitor::clean_names() |>
  dplyr::filter(!is.na(county)) |>
  dplyr::count(county, name = "n_schools")

ireland_counties <- ireland_counties |>
  dplyr::left_join(county_counts, by = c("NAME_1" = "county")) |>
  dplyr::mutate(n_schools = replace_na(n_schools, 0),
                hover_label = paste0(NAME_1, ": ", n_schools, " schools"))

leaflet(
  options = leafletOptions(zoomControl = TRUE, scrollWheelZoom = TRUE)
) |>
  addProviderTiles(providers$CartoDB.Positron) |>

  addPolygons(
    data = ireland_counties,
    fillOpacity = 0,
    color = "black",
    weight = 1,
    opacity = 0.8,
    label = ~hover_label,
    labelOptions = labelOptions(
      direction = "auto",
      textsize = "14px",
      style = list("font-weight" = "bold", "padding" = "6px 8px")
    ),
    highlightOptions = highlightOptions(
      weight = 3,
      color = "black",
      bringToFront = TRUE
    )
  ) |>

  addCircleMarkers(
    data = schools_sf,
    lng = ~school_longitude,
    lat = ~school_latitude,
    radius = radius,
    stroke = FALSE,
    fillOpacity = 0.65,
    color = "#a80050",
    popup = NULL
  ) |>

  fitBounds(
    lng1 = min(schools_sf$school_longitude, na.rm = TRUE),
    lat1 = min(schools_sf$school_latitude,  na.rm = TRUE),
    lng2 = max(schools_sf$school_longitude, na.rm = TRUE),
    lat2 = max(schools_sf$school_latitude,  na.rm = TRUE)
  )
```


## Datasets: Census Data

The census dataset comes from the **Central Statistics Office (CSO)** and provides demographic and socio-economic information at small-area level.

::: {.incremental}
- Includes population characteristics such as **age, sex, households, and families**.
- Provides socio-economic variables related to **education, social class, occupation, and commuting**.
- Contains information on **housing, car availability, and internet access**.
- Small-area data helps capture local demographic conditions around schools.
- These variables are used to explain how local context may influence school travel behaviour.
:::

## Datasets: Census



```{r active-transport-urban-rural}
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"

urb_rural_df <- read_csv("data/SAPS_2022_Small_Area_UR_171024.csv")%>%mutate(
    category = str_extract(UR_Category_Desc, "(?<=\\.).*"))%>%select("GEOGID","category", "T11_1_FSCCC","T11_1_BISCCC","T11_1_TSCCC","T11_1_NSSCCC","T11_1_WMFHSCCC")%>%distinct()%>%  mutate(
    urban_rural = case_when(
      category %in% c(
        " Cities",
        " Satellite Urban Towns",
        " Independent urban towns"
      ) ~ "Urban",
      
      category %in% c(
        " Rural areas with high urban influence",
        " Rural areas with moderate urban influence",
        " Highly rural/remote areas"
      ) ~ "Rural",
      
      TRUE ~ NA_character_
    )
  ) %>%
  rename(geogid = GEOGID)%>%
  mutate(
    denom = T11_1_TSCCC - T11_1_WMFHSCCC - T11_1_NSSCCC,
    perc_active = ifelse(
      denom > 0,
      100 * (T11_1_FSCCC + T11_1_BISCCC) / denom,
      NA_real_
    )
  )
plot_df <- urb_rural_df %>%
  filter(
    !is.na(perc_active),
    perc_active >= 0,
    perc_active <= 100,
    !is.na(urban_rural)
  )

p <- ggplot(plot_df, aes(x = perc_active)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 5,
    boundary = 0,
    closed = "left",
    alpha = 0.5
  ) +
  geom_density(
    aes(color = category),
    trim = TRUE,
    linewidth = 1,
    na.rm = TRUE
  ) +
  facet_wrap(~ urban_rural, nrow = 2) +
  labs(
    x = "Percentage using active transport",
    y = "Density",
    title = "Active Transport Distribution by Urban/Rural",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold")
  )

ggplotly(p) %>%
  layout(
    height = 480,
    margin = list(l = 60, r = 330, t = 70, b = 70),
    legend = list(
      orientation = "v",
      x = 1.03,
      y = 0.5,
      font = list(size = 10)
    )
  )
```
## Kriging

Kriging is a **spatial interpolation method** used to predict values at locations where we do not have observations.

::: {.incremental}
- It uses observed spatial data to estimate values at unsampled locations.
- Predictions are based on **nearby observations** and their spatial correlation.
- The spatial structure is described using a **variogram**.
- Points closer to the prediction location usually receive more weight.
- Kriging also provides a measure of **prediction uncertainty**.
:::

## Kriging

```{r}
```

## Travel Time

Travel time is used to measure **spatial accessibility**: how easily students can reach schools from their residential areas.

::: {.incremental}
- Travel times are calculated using a **routable transport network**.
- The method can include different modes, such as **walking, cycling, public transport, or driving**.
- Accessibility depends on the **origin**, the **destination**, and the available transport network.
- Time thresholds can be used to identify which schools are reachable within a given number of minutes.
- These measures help connect school location with students’ real mobility opportunities.
:::

## Travel time

PLOT

## Final Dataset

::: {.center}

```{mermaid}
%%{init: {"theme": "base", "themeVariables": {
  "primaryColor": "#f8eaf1",
  "primaryTextColor": "#222222",
  "primaryBorderColor": "#9f1f55",
  "lineColor": "#9f1f55",
  "fontSize": "26px"
}}}%%

flowchart LR
    A["School databases<br/><small>School location and attributes</small>"]
    B["Census<br/><small>Demographic variables</small>"]
    C["Kriging<br/><small>Spatial interpolation</small>"]
    D["Travel times<br/><small>Accessibility measure</small>"]
    E["Final dataframe<br/><small>Modelling dataset</small>"]

    A --> C
    B --> C
    C --> D
    A --> E
    B --> E
    D --> E

    classDef box fill:#f8eaf1,stroke:#9f1f55,stroke-width:4px,color:#222222,font-size:26px;
    class A,B,C,D,E box;
```

:::

## Descriptive Results

Scatter plot between response variable and travel time / kriging.

## Model

::: {.model-flow}
**Proportion of students using a travel mode**  
↓  
**Beta distribution**  
↓  
**Explained by distance, travel time, accessibility,  
and school context**
:::

::: {.incremental}
- Designed for continuous proportions between **0 and 1**.
- Uses a **logit link** to keep predicted values inside the valid range.
- Estimates both the relationship between variables and the uncertainty around those relationships.
- Well suited for comparing **urban and rural** school travel patterns.
:::

## Model: Implementation

```{r}
#| echo: true
#| eval: false

model_urban_krig <- brm(
  formula = formula_beta,
  data = df_urban_model,
  family = Beta(link = "logit"),
  prior = priors,
  chains = 4,
  iter = 4000,
  warmup = 1000,
  seed = 123,
  control = list(adapt_delta = 0.95)
)
```

::: {.incremental}
::: {.small-text}
- The formula is built dynamically using the selected predictors.
- Weakly informative priors regularise the model estimates.
- The model is fitted separately for the **urban dataset**.
- Four chains and 4000 iterations improve posterior estimation.
- `adapt_delta = 0.95` helps reduce sampling problems.
:::
:::
## Model: Urban Results

Content here.

## Model: Rural Results

Content here.

## Model: Improvements

Content here.

## Next Steps

Content here.

## Conclusions

Content here.

## {.center}

::: {.main-title}
Thank you for your attention!
:::