---
title: "Distance, Mobility, and Mode Choice : A Spatial Analysis of School Travel Behaviour in Ireland"
authors:
  - name: Sananthaa Jagadesan Sethamaraikannan
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland.
    email: sananthaajagadesan@gmail.com
  - name: Judith Beltran Lopez
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland
    email: j.beltranlopez1@universityofgalway.ie
abstract: |
  Active travel to school is shaped by the spatial relationship between where children live, where schools are located, and how easily schools can be reached. This study investigates whether school-place accessibility is associated with differences in the proportion of children walking or cycling to school across Ireland. Data from Census 2022, the Department of Education, and OpenStreetMap were combined to construct school accessibility measures for Census Small Areas. Two complementary accessibility measures were developed. First, a kriging-based measure was used to estimate spatial variation in school-place density across Census Small Areas. Second, a routing-based travel-time accessibility measure was used to estimate the number of school places reachable within walking and cycling time thresholds using the road and path network. Bayesian beta regression was then used to model the association between school accessibility and the proportion of children travelling to school by active modes, adjusting for demographic composition using isometric log-ratio transformed covariates. This approach allows the relationship between school accessibility and active school travel to be examined while accounting for spatial variation in local population characteristics. The results provide evidence on whether differences in school-place accessibility contribute to unequal opportunities for active school travel in Ireland.
keywords:
  - active travel
  - school accessibility
  - school-place density
  - kriging
  - travel-time accessibility
  - Census Small Areas
  - Bayesian beta regression
  - Ireland
bibliography: references.bib
biblio-style: unsrt
header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{booktabs}
  - \usepackage{array}
output: rticles::arxiv_article
---

# Introduction

Active travel to school, particularly walking and cycling, is an important part of everyday mobility for children and young people. It is shaped not only by individual or household characteristics, but also by the spatial organisation of schools, neighbourhoods, and transport networks. Children are more likely to walk or cycle when schools are located within a feasible distance and can be reached through safe and connected routes. For this reason, school accessibility is an important factor in understanding differences in travel behaviour to education.

In Ireland, access to nearby schools is not evenly distributed across space. Some areas have a high concentration of school places within short walking or cycling distances, while others have fewer nearby options. This has been described as a form of “postcode lottery”, where opportunities for active school travel may depend partly on where children live. However, this argument requires spatial evidence: it is necessary to measure school accessibility at a local scale and examine whether areas with better access to school places also have higher active travel rates.

This study investigates the relationship between school-place accessibility and active school travel across Ireland. The analysis combines Census 2022 Small Area data, Department of Education school enrolment and location data, and OpenStreetMap network data. Two complementary measures of school accessibility are developed. First, a kriging-based measure estimates the spatial distribution of school-place density across Census Small Areas. Second, a routing-based travel-time measure estimates how many school places can be reached within walking and cycling time thresholds using the road and path network.

The main outcome is the proportion of pupils travelling to school by active modes, defined as walking or cycling. Bayesian beta regression is used to model the association between school accessibility and active travel in urban Census Small Areas, while adjusting for demographic composition, car availability, and educational attainment. This allows the analysis to examine whether differences in school accessibility are associated with differences in active school travel after accounting for local population characteristics.

The objectives of this study are therefore to: construct local measures of school-place accessibility across Ireland; compare kriging-based and travel-time-based accessibility measures; estimate the association between school accessibility and active school travel; and assess whether spatial inequalities in school-place accessibility may contribute to unequal opportunities for walking and cycling to school.


## Background

School travel is not only a question of individual choice. It is also shaped by distance, school location, educational provision, and the structure of the local transport network. Walking and cycling to school are widely promoted because they can support healthier daily routines, reduce traffic congestion, and contribute to lower transport-related emissions. However, the opportunity to travel actively depends on whether schools are located within a feasible distance and whether they can be reached through suitable routes.

In Ireland, low levels of active travel to school are often discussed in terms of behaviour change, with policy attention directed towards encouraging pupils and families to walk or cycle. While behavioural factors are important, this explanation can overlook the spatial conditions that make active travel possible. If children do not have access to nearby schools, or if available schools are not reachable within reasonable walking or cycling times, then active travel may be difficult regardless of individual preferences.

Recent research has highlighted the importance of educational provision in shaping school travel patterns. Mancini argues that inequalities in school provision and school access form part of the wider Irish educational landscape [@mancini2024children]. Related work on adolescent active transport in Ireland suggests that active travel does not simply follow population size or density, but is strongly linked to the presence and type of local school provision [@mancini2025census]. This argument has also been discussed in a public-facing RTÉ Brainstorm article, which describes teenage active transport as a potential “postcode lottery” shaped by where young people live and what schools are available locally [@mancini2024rte].

This study builds on that research context by measuring school accessibility at a finer spatial scale. Instead of analysing only settlement-level differences, it uses Census Small Areas to examine local variation in school-place accessibility and active school travel. By combining school enrolment and location data, Census 2022 travel data, and OpenStreetMap network data, the study assesses whether areas with greater access to school places also have higher proportions of pupils walking or cycling to school.


## The Postcode Lottery Argument from the Article

The idea of a “postcode lottery” provides an important motivation for this study. In the context of school travel, the term suggests that opportunities for walking or cycling to school may depend partly on where a child lives. If school places are unevenly distributed, then children in some neighbourhoods may have nearby schools that are reachable by active modes, while children in other areas may need to travel longer distances or rely on motorised transport.

Mancini’s work argues that active travel to school should not be understood only as a matter of individual behaviour or family preference. Instead, it is also shaped by the geography of school provision. The RTÉ Brainstorm article summarising this research states that disparities between areas are not simply explained by population size, but are linked to differences in educational provision [@mancini2024rte]. This is important because it challenges the assumption that low levels of walking or cycling are mainly the result of unwillingness to travel actively. In some places, active travel may be less feasible because suitable school places are not located nearby or are not accessible through the local transport network.

This argument is also connected to wider research on inequalities in the Irish educational landscape. Mancini’s study of unequal school provision examines how the distribution and type of schools can create different educational opportunities across communities [@mancini2024children]. Related research on adolescent active transport in Ireland links school provision with unequal access to walking and cycling opportunities, showing that the spatial organisation of schools is relevant to sustainability, mobility, and education policy [@mancini2025census].

In this study, the “postcode lottery” argument is treated as an empirical question. Rather than assuming that school access directly determines active travel, the analysis tests whether areas with greater school-place accessibility also show higher proportions of pupils walking or cycling to school. This is done by constructing two complementary accessibility measures: a kriging-based measure of school-place density and a routing-based travel-time measure of reachable school places. By analysing these measures at Census Small Area level, the study examines whether spatial inequalities in school provision are associated with unequal opportunities for active school travel in Ireland.


## Research Objectives

Working with official statistics presents important methodological challenges for studying school travel behaviour at a detailed spatial scale. Census and administrative datasets are usually released in aggregated form, meaning that individual residential locations, school destinations, and travel routes are not directly observed. This study therefore combines spatial modelling, accessibility analysis, and Census Small Area data to examine how school-place accessibility relates to active school travel across Ireland.

The first objective is to map and analyse the spatial distribution of schools across Ireland, using school location and enrolment data to describe where school places are concentrated.

The second objective is to develop local measures of school accessibility. This includes a kriging-based measure of school-place density and a routing-based travel-time measure of the number of school places reachable by walking or cycling.

The third objective is to describe school travel behaviour using Census 2022 data, focusing on the prevalence of active travel modes, defined as walking and cycling, across Census Small Areas.

The fourth objective is to examine the relationship between school accessibility and active school travel in urban Census Small Areas, while adjusting for demographic composition, car availability, and educational attainment.

Together, these objectives aim to assess whether spatial differences in school-place accessibility are associated with unequal opportunities for walking and cycling to school in Ireland.


# Data

## School Data

School data were obtained from the Department of Education and Youth’s 2024/25 primary and post-primary school enrolment datasets [@departmenteducation_primary_2024_25; @departmenteducation_postprimary_2024_25]. These datasets provide school-level information for individual schools in Ireland, including school location coordinates and enrolment figures. The 2024/25 datasets were used because they provide recent enrolment information together with latitude and longitude coordinates for each school. These coordinates allowed schools to be represented directly as spatial point data without requiring additional geocoding.

The school datasets were used to construct the spatial inputs for both school accessibility measures used in this study. Each record represents an individual school and includes its geographic location and enrolment per return. Enrolment was used as a proxy for the number of school places available at each location. Although enrolment does not measure spare capacity directly, it provides a consistent measure of the relative size of schools and the spatial distribution of school places across the country.

The key variables extracted from the school datasets were school latitude, school longitude, and enrolment. School records without valid latitude or longitude coordinates were excluded because they could not be represented spatially. Enrolment values were checked before analysis to ensure that the school-place measure was based on valid pupil counts. No schools were removed for missing, zero, or invalid enrolment values after this checking step.

All school coordinates were initially provided in latitude and longitude using the WGS84 coordinate reference system. The school points were then transformed to the Irish Transverse Mercator projection, EPSG:2157. This projected coordinate system was used because distances, areas, and spatial operations are measured in metres, making it appropriate for the kriging analysis and for linking school locations to Census Small Areas.

Where multiple school records shared identical coordinates, enrolments were summed before analysis. This ensured that co-located schools, such as junior and senior schools on the same campus, contributed their combined enrolment to the school-place measure. Aggregating duplicate coordinates in this way avoided duplicated spatial locations while preserving the total number of pupils represented at each site.

The cleaned school point dataset was then used in two complementary parts of the accessibility analysis. First, the school locations and enrolment values were used as the observed input for the kriging-based school-place density measure. This produced an area-level estimate of school-place availability across Census Small Areas. Second, the same school locations were used as destination points in the routing-based travel-time accessibility analysis. In that analysis, school enrolment was used to weight destinations so that accessibility was measured as the number of school places reachable within specified walking and cycling travel-time thresholds, rather than simply the number of schools reachable.

This means that the school dataset contributed both a spatial density measure and a network-based travel-time measure of school accessibility. The kriging-based measure captures the broader spatial distribution of school places, while the routing-based measure captures how many school places can be reached through the actual road and path network within a given travel time. Together, these two measures allow school accessibility to be assessed from both a spatial-proximity perspective and a practical travel-time perspective.


## Census Data

Census data were obtained from the Central Statistics Office (CSO). The main census dataset used in the statistical modelling was the Census 2022 Small Area Population Statistics (SAPS) dataset [@cso_saps_2022]. Census Small Areas were used because they provide a fine spatial resolution and can be directly linked to the school accessibility measures constructed from school locations, kriging predictions, and routing-based travel-time calculations.

The main outcome variable was the proportion of pupils travelling to school by active modes. Active travel was defined as walking or cycling. The outcome was calculated from Census 2022 travel-to-education variables at Census Small Area level, excluding records with no stated mode of travel or where the question was not applicable. This produced an active-travel percentage for each Census Small Area that could be linked to the school accessibility measures and used as the response variable in the statistical model.

Census 2022 SAPS data were also used to derive demographic and population covariates. These included total population, population aged 3–18 years, age composition, household car availability, educational attainment, and urban–rural classification. The population aged 3–18 years was used to normalise the kriging-derived school-place measure, so that school-place availability was expressed relative to the number of school-age children living in each Census Small Area. Urban–rural classification was used to identify the urban Census Small Areas used in the main modelling analysis.

In addition to the Census Small Area data used in the model, settlement-level CSO tables were used for descriptive and contextual analysis of travel-to-education patterns. For 2022, table SAP2022T11T1TOWN22 provides travel-to-education and childcare data by mode for Towns/Urban Areas [@cso_sap2022t11t1town22]. For comparison with the previous census, table SAP2016T11T1SET provides equivalent 2016 settlement-level data for travel to education [@cso_sap2016t11t1set]. Table E6013 provides 2016 travel-to-education data separated by different periods of education for a smaller set of settlements [@cso_e6013]. These settlement-level tables were not used as the main modelling geography, but they provided broader context for interpreting active-travel patterns across Irish settlements.

Overall, the census data provided three key components for the analysis: the active-travel outcome, the demographic and population covariates, and the spatial units used to link travel behaviour with school accessibility. The Census Small Area data formed the basis of the main statistical modelling, while the settlement-level tables were used only to support descriptive comparisons and contextual interpretation.


## Study Area

The study area was the Republic of Ireland. The analysis was conducted at the Census Small Area level, using Census Small Area boundaries from the 2022 national statistical boundary dataset [@tailte_smallareas_2022]. Census Small Areas were selected as the main spatial unit because they provide a fine geographical scale for linking population characteristics, school accessibility, and travel behaviour. This scale is more appropriate for studying school travel than larger settlement-level geographies, because walking and cycling to school are highly local behaviours.

Each Census Small Area was treated as a neighbourhood-level unit of analysis. Census 2022 demographic and travel variables were linked to these boundaries, while school accessibility measures were derived from school locations and then summarised at the Census Small Area level. For the kriging-based measure, school-place density predictions were aggregated to Census Small Areas. For the routing-based travel-time measure, Census Small Areas were used as the origins from which reachable school places were calculated using the road and path network.

All spatial layers were transformed to the Irish Transverse Mercator coordinate reference system, EPSG:2157, before distance, area, and accessibility calculations were carried out. Using a projected coordinate system ensured that spatial measurements were calculated in metres and were consistent across the analysis.

The main statistical modelling was restricted to urban Census Small Areas. This restriction was used because the research question focuses on active school travel, and walking and cycling are more feasible in urban settings where population density, school density, and street-network connectivity are generally higher. Rural Census Small Areas were retained for descriptive spatial comparisons, but they were not included in the main Bayesian beta regression model.

# Methods

## School Place Density: Kriging

A kriging-based school-place density measure was developed to estimate spatial variation in school-place availability across Census Small Areas. Kriging is a geostatistical interpolation method that uses the spatial correlation structure of observed point data to predict values at unsampled locations. In this study, the observed point data were school locations, and the variable interpolated was school enrolment, used as a proxy for school-place availability. The general kriging workflow, including empirical variograms, fitted variogram models, and spatial prediction, follows standard geostatistical practice and is described in Moraga’s spatial statistics text [@moraga2024spatial; @moraga2024kriging]. The kriging implementation was carried out using the `gstat` package in R [@pebesma2004gstat].

School records were first converted to spatial points using their latitude and longitude coordinates and were then transformed to the Irish Transverse Mercator coordinate reference system, EPSG:2157. This ensured that distance-based calculations were carried out in metres. Records without valid coordinates or valid enrolment values were removed because they could not contribute to the spatial interpolation. Where multiple schools shared exactly the same coordinates, their enrolments were summed before analysis. This ensured that co-located schools, such as junior and senior schools on the same campus, contributed their combined enrolment to the school-place measure. After aggregation, the total enrolment at each unique school location was log-transformed before fitting the kriging model.

Ordinary kriging was used to predict school-place availability at unsampled locations. The spatial dependence in log-enrolment was first assessed using an empirical variogram. The empirical variogram shows how dissimilarity in log-enrolment changes as the distance between pairs of school locations increases. A fitted variogram model was then used to describe this spatial dependence formally and to provide the covariance structure required for kriging prediction. In the final analysis, a Gaussian variogram model was fitted to the empirical variogram. The variogram was estimated using a lag width of 5 km and a maximum cutoff distance of 200 km. Kriging predictions were made using up to 150 nearby observed school locations for each prediction point.

Instead of predicting only at the centroid of each Census Small Area, each Census Small Area polygon was divided into a 3 × 3 internal grid. Grid cells were clipped to the Census Small Area boundary, and one internal point was created within each valid clipped grid cell using a point-on-surface approach. Kriging predictions were then obtained at these internal prediction points and averaged to produce one raw kriging-derived estimate for each Census Small Area. This approach reduced dependence on a single centroid and better represented irregularly shaped areas.

Because the kriging model was fitted on the log-enrolment scale, predictions were transformed back to the original enrolment scale before aggregation. A bias-corrected log-normal back-transformation was used:

$$
\widehat{E}_i =
\exp \left(
\hat{z}_i + \frac{1}{2}\hat{\sigma}^{2}_i
\right)
$$

where (\hat{z}_i) is the kriging prediction on the log scale at prediction point (i), and (\hat{\sigma}^{2}_i) is the kriging prediction variance at that point. This correction was used because simply exponentiating the predicted log value would underestimate the expected value on the original enrolment scale.

The back-transformed predictions were averaged across the internal prediction points within each Census Small Area to produce one raw school-place estimate per area. The corresponding kriging prediction variance was also averaged across the prediction points to provide an area-level measure of interpolation uncertainty.

The raw kriging estimate represents predicted school-place availability, but it does not account for the number of children living in each area. To account for local demand, the raw kriged value was divided by the Census 2022 population aged 3–18 years in each Census Small Area. The final normalised school-place density measure was expressed per 100 children aged 3–18:

$$
SchoolDensity_i =
100 \times
\frac{RawKrigedValue_i}{PopulationAged3to18_i}
$$

where (RawKrigedValue_i) is the raw kriging-derived school-place estimate for Census Small Area (i), and (PopulationAged3to18_i) is the number of residents aged 3–18 in that area. Higher values indicate greater predicted school-place availability relative to the local school-age population. Lower values indicate lower predicted availability relative to potential local demand. This normalised measure was used as the main kriging-based school accessibility exposure in the statistical model. The unnormalised kriging estimate was retained as a reference measure of absolute predicted school-place availability, and school-place density per square kilometre was also calculated as a secondary spatial density measure.

### Kriging diagnostic figures

The following figures were used to check the spatial input data, the fitted variogram, and the resulting kriging-derived school-place density surface. These figures are included as diagnostics and descriptive outputs rather than as part of the model-fitting code.

```{r kriging-plot-setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(scales)
library(stringr)
library(gstat)

obj <- readRDS("../../outputs/combined_kriging_objects.rds")

dat_irl <- obj$dat_irl
variogram_emp <- obj$variogram
variogram_model <- obj$variogram_model
final_sf <- obj$final_sf
final_df <- obj$final_df
final_df_urban <- obj$final_df_urban

final_sf_plot <- final_sf %>%
  st_make_valid() %>%
  filter(
    is.finite(krig_value_raw),
    is.finite(krig_value_per_100_3_18),
    is.finite(pop_3_18)
  ) %>%
  mutate(
    krig_raw_cap = pmin(
      krig_value_raw,
      quantile(krig_value_raw, 0.99, na.rm = TRUE)
    ),
    krig_per100_cap = pmin(
      krig_value_per_100_3_18,
      quantile(krig_value_per_100_3_18, 0.99, na.rm = TRUE)
    )
  )

schools_plot <- dat_irl$schools_agg_sf %>%
  st_transform(st_crs(final_sf_plot)) %>%
  mutate(
    enrolment_plot = case_when(
      "enrolment_total" %in% names(.) ~ enrolment_total,
      "log_enrol" %in% names(.) ~ exp(log_enrol),
      TRUE ~ NA_real_
    )
  )

ireland_boundary <- dat_irl$region_sf %>%
  st_transform(st_crs(final_sf_plot)) %>%
  st_make_valid()

counties_shp_path <- "../../data/gadm41_IRL_shp/gadm41_IRL_2.shp"

counties_sf <- st_read(counties_shp_path, quiet = TRUE) %>%
  st_transform(st_crs(final_sf_plot)) %>%
  st_make_valid()

galway_boundary <- counties_sf %>%
  filter(
    str_detect(NAME_1, regex("Galway", ignore_case = TRUE)) |
      str_detect(NAME_2, regex("Galway", ignore_case = TRUE))
  ) %>%
  st_union() %>%
  st_as_sf() %>%
  mutate(county = "Galway")

galway_sf_plot <- final_sf_plot[galway_boundary, ]
schools_galway <- schools_plot[galway_boundary, ]
```

```{r school-locations-kriging-input, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of school locations used as input for the kriging analysis. Points represent aggregated school locations, with point size proportional to total enrolment at each location."}
ggplot() +
  geom_sf(
    data = ireland_boundary,
    fill = "grey97",
    colour = "grey60",
    linewidth = 0.2
  ) +
  geom_sf(
    data = schools_plot,
    aes(size = enrolment_plot),
    alpha = 0.55
  ) +
  scale_size_continuous(
    name = "Enrolment",
    range = c(0.2, 2.2),
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "School locations used for kriging",
    subtitle = "Aggregated school locations weighted by total enrolment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r kriging-variogram-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Empirical and fitted variogram for log-enrolment. The x-axis shows distance between pairs of school locations, and the y-axis shows semivariance in log-enrolment."}
plot(
  variogram_emp,
  variogram_model,
  main = "Empirical and fitted variogram for log-enrolment"
)
```

```{r normalised-kriging-ireland-map, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Normalised kriged school-place density by Census Small Area. Values are expressed per 100 children aged 3–18. The colour scale is capped at the 99th percentile to make spatial patterns easier to see."}
ggplot() +
  geom_sf(
    data = final_sf_plot,
    aes(fill = krig_per100_cap),
    colour = NA
  ) +
  geom_sf(
    data = ireland_boundary,
    fill = NA,
    colour = "grey40",
    linewidth = 0.15
  ) +
  scale_fill_viridis_c(
    name = "School-place density\nper 100 children",
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "Normalised kriged school-place density",
    subtitle = "Per 100 children aged 3–18; colour capped at 99th percentile"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r kriging-density-distribution, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Distribution of the normalised kriging-based school-place density measure. The x-axis is limited at the 99th percentile to reduce the influence of extreme values in the plot."}
final_df %>%
  filter(is.finite(krig_value_per_100_3_18)) %>%
  ggplot(aes(x = krig_value_per_100_3_18)) +
  geom_histogram(bins = 40) +
  coord_cartesian(
    xlim = c(
      0,
      quantile(final_df$krig_value_per_100_3_18, 0.99, na.rm = TRUE)
    )
  ) +
  labs(
    title = "Distribution of kriged school-place density",
    subtitle = "Per 100 children aged 3–18; x-axis capped at 99th percentile",
    x = "School-place density per 100 children aged 3–18",
    y = "Number of Census Small Areas"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold")
  )
```

```{r galway-kriging-school-overlay, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Example local view of the normalised kriging-based school-place density measure in Galway. School points are overlaid to show how the interpolated surface relates to observed school locations."}
ggplot() +
  geom_sf(
    data = galway_sf_plot,
    aes(fill = krig_per100_cap),
    colour = NA
  ) +
  geom_sf(
    data = galway_boundary,
    fill = NA,
    colour = "black",
    linewidth = 0.4
  ) +
  geom_sf(
    data = schools_galway,
    aes(size = enrolment_plot),
    alpha = 0.7
  ) +
  scale_fill_viridis_c(
    name = "School-place density\nper 100 children",
    labels = label_number(big.mark = ",")
  ) +
  scale_size_continuous(
    name = "School enrolment",
    range = c(0.5, 3),
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "Galway kriging values with school locations",
    subtitle = "Normalised kriging per 100 children aged 3–18"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

The national map shows the broad spatial pattern in the kriging-derived school-place density measure, while the Galway map provides a clearer local example of how observed school locations contribute to the interpolated surface. The distribution plot is included because the normalised kriging variable is highly skewed; for this reason, later descriptive plots and maps use percentile capping where appropriate to improve readability.

### Normalisation by School-Age Population

The raw kriging-derived estimate represents predicted school-place availability, but it does not account for the number of children living in each Census Small Area. This adjustment was necessary because an area with a high predicted number of school places may not necessarily have high accessibility if it also has a large school-age population. To account for local demand, the raw kriged school-place estimate was normalised by the Census 2022 population aged 3–18 years in each Census Small Area [@cso_saps_2022].

The population aged 3–18 years was used because it approximates the population potentially requiring access to primary and post-primary education. Although this age range does not correspond exactly to school attendance in every case, it provides a consistent local denominator for comparing school-place availability across Census Small Areas.

The normalised kriging-based accessibility measure was calculated as:

$$
SchoolDensity_i =
100 \times
\frac{RawKrigedValue_i}{PopulationAged3to18_i}
$$

where (RawKrigedValue_i) is the raw kriging-derived school-place estimate for Census Small Area (i), and (PopulationAged3to18_i) is the number of residents aged 3–18 in that area. Multiplying by 100 expresses the measure as predicted school-place availability per 100 children aged 3–18, which is easier to interpret at the Census Small Area scale than a per-child value.

Higher values indicate greater predicted school-place availability relative to the local school-age population. Lower values indicate lower predicted availability relative to potential local demand. This means that the normalised measure should be interpreted as a relative accessibility measure, rather than as the exact number of school places physically located within each Census Small Area.

The unnormalised kriging value was retained as a reference measure of absolute predicted school-place availability. An additional spatial density variable, predicted school places per square kilometre, was also calculated to describe the concentration of school places independently of the local school-age population denominator. The normalised measure per 100 children aged 3–18 was used as the main kriging-based accessibility exposure in the Bayesian beta regression model.


## Travel Time Accessibility:r5r

## Demographic Variables: ILR Transformation

Several demographic variables derived from Census 2022 Small Area Population Statistics were included to adjust for differences in population structure across Census Small Areas. These variables describe age structure, household car availability, and educational attainment. Because these variables are expressed as percentages within groups, they are compositional: the parts represent shares of a whole and are therefore constrained to sum to a fixed total. Standard regression models can give unstable or misleading results when compositional variables are included directly, because an increase in one component necessarily implies a decrease in at least one other component [@aitchison1986compositional].

Three demographic compositions were constructed. Age composition was grouped into five categories: children aged 0–12, young people aged 13–24, adults aged 25–44, adults aged 45–74, and adults aged 75 and over. Household car availability was grouped into households with no car, one car, and two or more cars. Educational attainment was grouped into no formal or primary education, secondary education, and tertiary education. The percentage still in education was retained as a separate demographic covariate, rather than being included in the educational attainment composition, because it was calculated using a different denominator.

To account for the compositional nature of these variables, the isometric log-ratio transformation was applied to each composition. The ILR transformation converts a composition with (D) parts into (D - 1) real-valued coordinates that can be used in standard regression models [@egozcue2003ilr]. These coordinates represent balances between groups of compositional parts rather than isolated changes in individual percentages. For example, an ILR coordinate for age composition may represent a contrast between younger and older age groups, rather than the independent effect of one age group alone.

For a composition (\mathbf{x}*i = (x*{i1}, x_{i2}, ..., x_{iD})) in Census Small Area (i), the transformation can be written generally as:

$$
\mathbf{z}_i = ilr(\mathbf{x}_i)
$$

where (\mathbf{z}_i) is the set of (D - 1) ILR coordinates. These transformed coordinates are unconstrained and can therefore be included as regression covariates without the perfect collinearity that would arise from including all raw percentage components directly.

Exact zero values cannot be used directly in log-ratio transformations because the logarithm of zero is undefined. Before applying the ILR transformation, exact zero percentages were therefore replaced with a small positive value of 0.001. This allowed the transformation to be computed while making only a minimal change to the original percentage values.

The ILR transformation was implemented in R using the `compositions` package [@vandenboogaart2008compositions]. The resulting ILR-transformed demographic variables were included as covariates in the Bayesian beta regression model. This allowed the model to adjust for demographic differences between urban Census Small Areas while avoiding the collinearity problems associated with raw percentage variables. The original percentage variables were retained for descriptive summaries, but the ILR-transformed variables were used in the regression model.



## Statistical Model: Bayesian Beta Regression

The main statistical model was used to estimate the association between school accessibility and active school travel in urban Census Small Areas. The outcome variable was the proportion of pupils travelling to school by active modes, defined as walking or cycling. Because this outcome is continuous and bounded between 0 and 1, beta regression was used. Beta regression is appropriate for modelling rates and proportions that lie within the open interval ((0, 1)), and allows the expected value of the outcome to be related to explanatory variables through a regression structure [@ferrari2004beta].

The analysis was restricted to urban Census Small Areas because the main research question focuses on active school travel in settings where walking and cycling are more feasible. Before modelling, Census Small Areas with missing required model inputs were removed. In addition, areas with total population greater than 900 and areas with fewer than five residents aged 3–18 were excluded. These exclusions were applied to reduce the influence of unusually large Census Small Areas and unstable school-age population denominators. The final urban modelling dataset therefore included Census Small Areas with valid active-travel outcomes, valid school accessibility measures, and sufficient population information.

The active-travel percentage was converted to a proportion before modelling. Since the beta distribution is defined only on the open interval ((0, 1)), any exact zero or one values were adjusted slightly so that all observations lay strictly within this interval. This type of transformation is commonly used when applying beta regression to proportional outcomes that may include boundary values [@smithson2006better].

The model was fitted using a Bayesian beta regression framework. For Census Small Area (i), the outcome was modelled as:

$$
y_i \sim \mathrm{Beta}(\mu_i, \phi)
$$

where (y_i) is the active-travel proportion, (\mu_i) is the expected active-travel proportion, and (\phi) is the beta precision parameter. Under this parameterisation, (E(y_i) = \mu_i). Larger values of (\phi) indicate that observations are more tightly concentrated around the expected proportion, while smaller values indicate greater variability.

A logit link function was used to relate the expected active-travel proportion to the explanatory variables:

$$
\mathrm{logit}(\mu_i) =
\alpha +
\sum_{j=1}^{p} \beta_j Z_{ij}
+
f_1(K_i)
+
f_2(D_i)
$$

where (\alpha) is the intercept, (Z_{ij}) represents the set of standardised demographic covariates for Census Small Area (i), (K_i) is the normalised kriging-based school-place density measure, and (D_i) is the secondary spatial density measure of predicted school places per square kilometre. The demographic covariates included the ILR-transformed age composition, household car-availability composition, and educational-attainment composition, together with the additional percentage still in education covariate where included. These variables were used to adjust for differences in demographic structure across urban Census Small Areas.

The terms (f_1(K_i)) and (f_2(D_i)) represent natural cubic spline functions. Natural cubic splines were used because the relationship between school accessibility and active travel may not be linear. For example, increases in school accessibility may be strongly associated with active travel at low levels of accessibility, but the association may become weaker or flatten at higher levels. Using splines allows the model to capture this type of non-linear relationship without assuming a single constant effect across the full range of the accessibility variables. Three degrees of freedom were used for each spline, providing a flexible but controlled curve and reducing the risk of overfitting.

The normalised kriging-based accessibility variable was expressed as predicted school-place availability per 100 children aged 3–18. This variable captures school-place availability relative to local school-age population demand. The additional school-place density per square kilometre variable was included to capture the spatial concentration of predicted school places independently of the local child population denominator. Including both variables allowed the model to distinguish between relative school-place availability and broader spatial concentration of school places.

To account for differences in the reliability of active-travel percentages across Census Small Areas, the model used weights based on the active-travel denominator. Census Small Areas with larger numbers of pupils with a stated travel mode provide more stable estimates of active-travel proportion than areas with smaller denominators. The weights were calculated as the square root of the active-travel denominator and then normalised to have an average value of one in the urban modelling sample. This gave more influence to areas with more reliable active-travel estimates while avoiding excessive dominance by the largest areas.

Weakly informative priors were used for the model parameters. Regression coefficients were assigned normal priors centred at zero, which regularised the model by discouraging unrealistically large effects while still allowing the data to inform the posterior estimates. The intercept was assigned a wider normal prior on the logit scale. The beta precision parameter (\phi) was assigned a positive prior to ensure that it remained greater than zero. These priors were chosen to provide regularisation without imposing strong assumptions about the direction or size of the accessibility effects.

The model was fitted using the `brms` package in R, which fits Bayesian regression models using Stan [@burkner2017brms]. Posterior sampling was carried out using Hamiltonian Monte Carlo with the No-U-Turn Sampler, as implemented in Stan [@hoffman2014nuts]. Four Markov chains were run, and convergence was assessed using standard Bayesian diagnostics, including (\hat{R}), effective sample size, trace plots, divergent transitions, and posterior predictive checks. Values of (\hat{R}) close to 1 and adequate effective sample sizes were taken as evidence that the chains had mixed well [@vehtari2021rhat].

Because spline coefficients are not directly interpretable as simple one-unit changes, the results were interpreted using posterior predictions and marginal effect plots on the response scale. These plots show how the expected active-travel percentage changes across the observed range of the accessibility variables. Interpretation was based on posterior means and credible intervals, rather than only on raw regression coefficients. Posterior predictive checks were used to assess whether the fitted model could reproduce the observed distribution of active travel.

# Results

## Spatial Distribution of Schools

The 2024/25 school dataset was first explored spatially to understand the distribution of school locations across Ireland. Each school was represented as a point using its latitude and longitude coordinates, and points were mapped against the national boundary. This provides an initial descriptive view of where schools are concentrated before constructing the kriging-based and travel-time accessibility measures.

```{r spatial-distribution-schools, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of schools across Ireland in the 2024/25 mainstream school dataset. Points represent school locations and are coloured by school level."}
library(readxl)
library(dplyr)
library(sf)
library(ggplot2)
library(rnaturalearth)

schools_df <- read_excel(
  "../data/Data_on_Individual_Schools_Mainstream_2024_25.xlsx",
  sheet = 2
) %>%
  rename(
    Roll_Number = `Roll Number`,
    longitude = `School Longitude`,
    latitude = `School Latitude`,
    AcademicYear = `Academic Year (Enrolment)`,
    SchoolName = `Official Name`
  )

schools_df_clean <- schools_df %>%
  filter(!is.na(latitude), !is.na(longitude))

schools_sf <- st_as_sf(
  schools_df_clean,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
)

ireland <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Ireland")

ggplot() +
  geom_sf(
    data = ireland,
    fill = "grey95",
    colour = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data = schools_sf,
    aes(colour = `School Level`),
    size = 0.7,
    alpha = 0.75
  ) +
  labs(
    title = "Spatial distribution of schools in Ireland",
    subtitle = "Mainstream schools, 2024/25",
    colour = "School level"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

The map shows that schools are distributed across the country, with visible clustering in and around larger urban areas. This spatial pattern is important for the later analysis because school accessibility is not determined only by the number of schools nationally, but by how school places are distributed relative to where children live. The mapped school locations therefore provide the spatial input for both the kriging-based school-place density measure and the routing-based travel-time accessibility analysis.


## Walking and Cycling Accessibility: travel time

## Kriging vs Travel Time Comparison

## Model Results — Urban Areas

## Effect of Walking Accessibility on Active Travel

# Discussion

## Interpretation of Results

## Limitations

This study has several limitations. First, the outcome data are restricted by the way school travel information is released in official statistics. The Census data provide aggregated travel-to-education counts for Census Small Areas, rather than individual-level journeys between home and school. As a result, the analysis cannot observe the actual school attended by each pupil, the exact route taken, or whether pupils travel to primary or post-primary schools. The model therefore estimates area-level associations between school accessibility and active travel, rather than individual travel behaviour. This means that the results should be interpreted as evidence of spatial association, not as proof that improved accessibility directly causes higher walking or cycling rates.

Second, the analysis combines school accessibility measures with an outcome that is not fully separated by school type in the main Small Area-level data. Primary and post-primary pupils may have different travel distances, school-choice patterns, and active travel opportunities. For example, younger children may be more dependent on nearby primary schools, while secondary students may travel further because post-primary schools are fewer and more specialised. Future work should therefore repeat the analysis separately for primary and post-primary schools where the data allow. However, this may be limited by the availability of travel-mode outcomes disaggregated by school type at fine spatial scales.

Third, the school accessibility measures use enrolment as a proxy for school places. Enrolment provides a consistent measure of the relative size and spatial distribution of schools, but it does not measure spare capacity, catchment areas, admissions policies, or whether a place is realistically available to a child living in a particular area. This limitation is especially relevant where schools are oversubscribed or where school choice is shaped by factors other than distance.

Fourth, the kriging-based accessibility measure captures the spatial distribution of school places, but it does not account for barriers in the transport network, route safety, traffic conditions, or the quality of walking and cycling infrastructure. The routing-based travel-time measure addresses this limitation more directly by using the road and path network, but it still depends on assumptions about walking and cycling speeds and on the completeness and accuracy of OpenStreetMap data.

Finally, the workflow could be repeated using settlement-level data as a complementary analysis. Settlement-level travel tables can describe travel behaviour for whole towns or urban areas, rather than only Census Small Areas. This may provide a broader view of how school provision and active travel relate at the town scale, and could be useful for comparing with previous work on Irish settlements. However, settlement-level data are less spatially detailed than Census Small Area data, so they would be better suited as a complementary robustness or contextual analysis rather than a replacement for the main local-scale model.


## Comparison with the article

# Conclusions
