---
title: "Distance, Mobility, and Mode Choice : A Spatial Analysis of School Travel Behaviour in Ireland"
authors:
  - name: Sananthaa Jagadesan Sethamaraikannan
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland.
    email: sananthaajagadesan@gmail.com
  - name: Judith Beltran Lopez
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland
    email: j.beltranlopez1@universityofgalway.ie
abstract: |
  In Ireland, whether a child can walk or cycle to school depends heavily on where they live, something that has been described as a postcode lottery. This study investigates whether the location of schools and how easily they can be reached explains differences in active travel rates across Ireland. We used data from the 2022 Census, the Department of Education, and OpenStreetMap to measure school accessibility for over 16,000 small areas across Ireland. Two methods were used to measure accessibility-kriging, which estimates the density of school places across space, and r5r, which calculates how many school places can be reached within a given walking or cycling time using the real road network. We then used Bayesian Beta regression to model the relationship between accessibility and the proportion of children walking or cycling to school.
keywords:
  -sss
bibliography: references.bib
biblio-style: unsrt
header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{booktabs}
  - \usepackage{array}
output: rticles::arxiv_article
---

# Introduction

Here goes an introduction text

## 1.1 Background

## 1.2 The Postcode Lottery Argument from the article


## 1.3 Research Objectives

# 2 Data

## 2.1 School Data

The school data used in this study were obtained from the Department of Education and Youth’s *Data on Individual Schools* collection. This source provides school-level information for individual schools in Ireland, including enrolment figures. For this analysis, the 2024/25 mainstream school dataset was used because it provides recent school-level enrolment information together with school location coordinates.

The school dataset was used to construct the main spatial input for the school accessibility analysis. Each record represents an individual school and includes information such as school location, school latitude and longitude, and enrolment. In this study, enrolment was used as a proxy for the number of available school places. The geographic coordinates were used to convert each school into a spatial point, allowing the school data to be linked with Small Area boundaries and used in the kriging analysis.

The key variables extracted from the school dataset were:

* school latitude;
* school longitude;
* enrolment per return.

Schools with missing latitude or longitude were excluded because they could not be represented spatially. Schools with missing, zero, or invalid enrolment values were also excluded because enrolment was required for the school-place accessibility measure. Enrolment was then log-transformed before spatial interpolation, as the raw enrolment distribution was positively skewed and contained some very large schools.

All school coordinates were initially provided as latitude and longitude and were converted into spatial point data using the WGS84 coordinate reference system. The points were then transformed to the Irish Transverse Mercator projection, EPSG:2157, so that spatial distances and areas could be calculated in metres. This projected coordinate system was used consistently throughout the spatial analysis.

Where multiple school records shared identical coordinates, these points were aggregated by taking the mean log-enrolment value. This avoided duplicated spatial locations in the kriging model and ensured that each coordinate pair contributed only one observed value to the spatial interpolation.

The cleaned school point dataset was used as the observed input for the ordinary kriging analysis described in Section 3.1. The aim of this step was not to count the number of schools within each Small Area directly, but to estimate a continuous spatial surface of school-place intensity across Ireland. This kriging-derived surface was later aggregated to Census Small Areas and normalised by the population aged 3–18 to produce the final school place density variable.


## 2.2 Census Data

Census data were obtained from the Central Statistics Office (CSO), using the Small Area Population Statistics (SAPS) and settlement-level tables from Census 2022 and Census 2016. These data were used to describe travel behaviour to education and to provide demographic and spatial context for the analysis.

The main census outcome of interest was travel mode to education. Settlement-level aggregated tables were used to obtain the proportion of people travelling to education or childcare by different modes of travel. For Census 2022, the relevant table was `SAP2022T11T1TOWN22`, which provides travel-to-education data for Towns/Urban Areas. This table combines travel to all forms of education and childcare and covers approximately 800 settlements. For comparison with the previous census, the 2016 table `SAP2016T11T1SET` provides equivalent settlement-level data for all forms of education combined. An additional 2016 table, `E6013`, provides travel-to-education data separated by different stages or periods of education, but for a smaller set of settlements.

These tables were accessed through the CSO’s PxStat database. They can be downloaded manually from the CSO website or programmatically in R using the `csodata` package. Accessing the tables through R ensures that the data extraction is reproducible and that the table codes used in the analysis are clearly documented.

The settlement-level census data were linked to spatial boundary data using CSO Towns/Urban Areas boundaries. For the 2022 analysis, the relevant spatial geography is the CSO Towns/Urban Areas 2022 layer. These boundaries correspond to the Built Up Areas geography used in Census 2022. Built Up Areas were introduced as a new urban geography and are based on clusters of urban development. Their boundaries are designed to be compatible with Census 2022 Small Areas, which supports consistent spatial analysis.

The census travel tables were used to calculate the proportion of people travelling to education by active modes. Active travel was defined as walking or cycling. The active travel percentage was calculated as:

[
\text{Active travel percentage} =
100 \times
\frac{
\text{Number travelling by foot} + \text{Number travelling by bicycle}
}{
\text{Total travelling to education, excluding not stated and not applicable categories}
}
]

This measure was used to describe variation in active travel to education across settlements and to provide contextual evidence for the later Small Area-level modelling. The settlement-level data were particularly useful for understanding broader urban patterns, while the Small Area SAPS data were used in the main statistical modelling because they provide a finer spatial resolution and can be linked directly to the school accessibility measures.

In addition to travel mode, Census 2022 SAPS data were used to derive demographic covariates for the statistical model. These included age composition, car availability, educational attainment, population aged 3–18 years, total population, and urban–rural classification. The population aged 3–18 years was used to normalise the kriging-derived school place measure, producing a school place density variable per 1,000 children. Urban–rural classification was used to restrict the main modelling analysis to urban Small Areas, which were the focus population of the study.

Overall, the census data provided three key components for the analysis: the active travel outcome, demographic control variables, and spatial boundaries for linking statistical data to geographic areas.


## 2.3 Study Area

# 3 Methods

## 3.1 School Place Density: Kriging

This section describes the construction of the school place density variable used as the main spatial accessibility exposure in the analysis. The aim was to estimate how the availability of school places varies spatially across Ireland and to derive a Small Area-level measure that could be linked to Census 2022 demographic and travel behaviour data.

The input school data consisted of individual mainstream schools for the 2024/25 academic year, including school coordinates and enrolment counts. Since school enrolment was used as a proxy for school place availability, schools with missing coordinates or invalid enrolment values were removed. Enrolment was log-transformed before interpolation because the original enrolment distribution was positively skewed.

All spatial datasets were transformed to the Irish Transverse Mercator coordinate reference system, EPSG:2157, so that distances and areas were measured consistently in metres. Schools with identical coordinates were aggregated by taking the mean log-enrolment value, avoiding duplicated point locations in the geostatistical model.

Ordinary kriging was used to predict school place availability at unsampled locations. Kriging is appropriate here because it uses the spatial autocorrelation structure of the observed school data to estimate values at locations where no school is directly observed. The spatial dependence in log-enrolment was first assessed using an empirical variogram. An exponential variogram model was then fitted and used to generate kriging predictions.

Instead of predicting only at the centroid of each Census Small Area, each Small Area was divided into a 3 × 3 internal grid. Prediction points were created within the valid clipped grid cells using a point-on-surface approach. Kriging predictions were obtained at these internal points and then averaged to obtain a single predicted value for each Small Area. This approach was used to better represent irregularly shaped Small Areas and reduce dependence on a single centroid.

The raw kriging output was back-transformed from the log scale using the exponential function. This produced a raw spatially interpolated estimate of school place intensity. However, because Small Areas differ in the number of children living in them, this raw value was normalised by the Census 2022 population aged 3–18 years. The final school place density variable was expressed as predicted school place intensity per 1,000 children aged 3–18:

[
\text{School place density}*{i} =
1000 \times
\frac{
\text{Raw kriged school place value}*{i}
}{
\text{Population aged 3--18}_{i}
}
]

Higher values indicate greater predicted school place availability relative to the local school-age population. Lower values indicate lower predicted availability relative to potential local demand. This normalised variable was used in the later statistical modelling as the main school place accessibility exposure.

```{r setup-kriging-section, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(gstat)
library(knitr)

# Load objects created by the kriging script
obj <- readRDS("outputs/combined_kriging_objects.rds")

dat_irl          <- obj$dat_irl
variogram_emp    <- obj$variogram
variogram_model  <- obj$variogram_model
kriged_points_sf <- obj$kriged_points_sf
final_sf         <- obj$final_sf
final_df         <- obj$final_df
final_df_urban   <- obj$final_df_urban

schools_sf <- dat_irl$schools_agg_sf %>%
  mutate(enrolment_est = exp(log_enrol))

final_sf_plot <- final_sf %>%
  filter(
    !is.na(urban_rural),
    is.finite(krig_value_raw),
    is.finite(krig_value_per_1000_3_18),
    is.finite(krig_var)
  )
```

### Observed school locations

```{r school-locations-map, fig.cap="Spatial distribution of mainstream schools used in the kriging analysis. Points represent school locations and are coloured by estimated enrolment. These observed locations form the input data for the interpolation.", message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = final_sf_plot, fill = "grey95", colour = NA) +
  geom_sf(
    data = schools_sf,
    aes(colour = enrolment_est),
    size = 0.8,
    alpha = 0.8
  ) +
  scale_colour_viridis_c(name = "Enrolment") +
  labs(
    title = "Observed school locations",
    subtitle = "School points used as input for kriging"
  ) +
  theme_minimal()
```

### Distribution of enrolment before and after log transformation

```{r enrolment-histogram, fig.cap="Distribution of school enrolment before log transformation. The raw enrolment variable is positively skewed, motivating the use of log-enrolment in the kriging model.", message=FALSE, warning=FALSE}
schools_sf %>%
  st_drop_geometry() %>%
  ggplot(aes(x = enrolment_est)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of raw school enrolment",
    x = "School enrolment",
    y = "Number of schools"
  ) +
  theme_minimal()
```

```{r log-enrolment-histogram, fig.cap="Distribution of log-transformed school enrolment. The log transformation reduces skewness and limits the influence of very large schools in the spatial interpolation.", message=FALSE, warning=FALSE}
schools_sf %>%
  st_drop_geometry() %>%
  ggplot(aes(x = log_enrol)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of log school enrolment",
    x = "Log school enrolment",
    y = "Number of schools"
  ) +
  theme_minimal()
```

### Empirical and fitted variogram

The empirical variogram was used to assess whether school enrolment values were spatially autocorrelated. The fitted exponential variogram model was then used to derive the kriging weights.

```{r variogram-plot, fig.cap="Empirical variogram and fitted exponential variogram model for log school enrolment. The fitted model describes the spatial dependence structure used by ordinary kriging.", message=FALSE, warning=FALSE}
plot(
  variogram_emp,
  variogram_model,
  main = "Empirical and fitted variogram for log school enrolment"
)
```

### Small Area prediction points

```{r prediction-points-example, fig.cap="Example of internal prediction points within one Census Small Area. Each Small Area was divided into a 3 × 3 grid, clipped to the boundary, and represented using internal point-on-surface prediction locations.", message=FALSE, warning=FALSE}
example_sa_id <- kriged_points_sf %>%
  st_drop_geometry() %>%
  count(SA_PUB2022, name = "n_points") %>%
  arrange(desc(n_points)) %>%
  slice(1) %>%
  pull(SA_PUB2022)

example_sa <- final_sf %>%
  filter(SA_PUB2022 == example_sa_id)

example_pts <- kriged_points_sf %>%
  filter(SA_PUB2022 == example_sa_id)

ggplot() +
  geom_sf(data = example_sa, fill = "grey95", colour = "black") +
  geom_sf(data = example_pts, size = 2) +
  labs(
    title = "Example of 3 × 3 Small Area prediction points",
    subtitle = paste("Small Area:", example_sa_id)
  ) +
  theme_minimal()
```

### Raw kriged school place intensity

```{r raw-kriging-map, fig.cap="Raw kriged school place intensity by Small Area. Values represent the average back-transformed kriging prediction across the internal prediction points in each Small Area.", message=FALSE, warning=FALSE}
ggplot(final_sf_plot) +
  geom_sf(aes(fill = krig_value_raw), colour = NA) +
  scale_fill_viridis_c(name = "Raw kriged\nvalue") +
  labs(
    title = "Raw kriged school place intensity",
    subtitle = "Back-transformed prediction from the log-enrolment kriging model"
  ) +
  theme_minimal()
```

### Normalised school place density per 1,000 children aged 3–18

```{r normalised-kriging-map, fig.cap="Normalised school place density per 1,000 children aged 3–18 by Small Area. This is the main school accessibility exposure used in the modelling stage.", message=FALSE, warning=FALSE}
ggplot(final_sf_plot) +
  geom_sf(aes(fill = krig_value_per_1000_3_18), colour = NA) +
  scale_fill_viridis_c(name = "School place density\nper 1,000 children") +
  labs(
    title = "School place density per 1,000 children aged 3–18",
    subtitle = "Raw kriged value normalised by local school-age population"
  ) +
  theme_minimal()
```

### Kriging prediction variance

```{r kriging-variance-map, fig.cap="Kriging prediction variance by Small Area. Higher values indicate areas where the kriging prediction is more uncertain.", message=FALSE, warning=FALSE}
ggplot(final_sf_plot) +
  geom_sf(aes(fill = krig_var), colour = NA) +
  scale_fill_viridis_c(name = "Prediction\nvariance") +
  labs(
    title = "Kriging prediction variance",
    subtitle = "Uncertainty associated with the kriging predictions"
  ) +
  theme_minimal()
```

### Urban and rural comparison

```{r urban-rural-density-boxplot, fig.cap="Distribution of normalised school place density by urban-rural classification. This plot compares the school accessibility exposure across urban and rural Small Areas.", message=FALSE, warning=FALSE}
final_df %>%
  filter(
    !is.na(urban_rural),
    is.finite(krig_value_per_1000_3_18)
  ) %>%
  ggplot(aes(x = urban_rural, y = krig_value_per_1000_3_18)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "School place density by urban-rural classification",
    x = "Urban-rural classification",
    y = "School place density per 1,000 children aged 3–18"
  ) +
  theme_minimal()
```

### Relationship with active school travel in urban areas

```{r active-travel-kriging-scatter, fig.cap="Relationship between active school travel and normalised school place density in urban Small Areas. Each point represents one urban Small Area.", message=FALSE, warning=FALSE}
final_df_urban %>%
  filter(
    is.finite(perc_active),
    is.finite(krig_value_per_1000_3_18)
  ) %>%
  ggplot(aes(x = krig_value_per_1000_3_18, y = perc_active)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Active school travel and school place density",
    subtitle = "Urban Small Areas only",
    x = "School place density per 1,000 children aged 3–18",
    y = "Active school travel (%)"
  ) +
  theme_minimal()
```

### Final kriging dataset checks

```{r kriging-summary-checks, message=FALSE, warning=FALSE}
final_df %>%
  summarise(
    n = n(),
    missing_krig = sum(is.na(krig_value)),
    missing_perc_active = sum(is.na(perc_active)),
    min_krig = min(krig_value, na.rm = TRUE),
    max_krig = max(krig_value, na.rm = TRUE),
    min_krig_norm = min(krig_value_per_1000_3_18, na.rm = TRUE),
    max_krig_norm = max(krig_value_per_1000_3_18, na.rm = TRUE)
  ) %>%
  kable(
    digits = 3,
    caption = "Summary checks for the final kriging-derived dataset."
  )
```

The resulting dataset contains the raw kriged school place estimate, the normalised school place density per 1,000 children aged 3–18, the kriging prediction variance, and the Census-derived variables used in the later modelling stage. For interpretation, the normalised variable is preferred because it accounts for differences in local school-age population size. The raw kriged value is retained only as a reference measure of interpolated school place intensity.


## 3.2 Travel Time Accessibility:r5r

## 3.3 Demographic Variables: ILR Transformation

Several demographic covariates used in this study are expressed as percentages. These include age composition, household car availability, and educational attainment. Although percentages are easy to interpret descriptively, they require special treatment in regression modelling because they are compositional variables. A composition is a set of parts that together represent a whole. For example, the age-group percentages in a Small Area represent the distribution of the local population across age categories, while the car-ownership percentages represent the distribution of households according to the number of cars available.

The main issue with compositional variables is that their components are not statistically independent. If one percentage increases, at least one other percentage must decrease because the parts are constrained by a fixed total. This means that raw percentages do not behave like ordinary unconstrained regression predictors. Including all components directly in a regression model can lead to perfect or near-perfect collinearity, unstable coefficients, and misleading interpretations.

To address this issue, the demographic percentage variables were transformed using the isometric log-ratio transformation. The ILR transformation converts a composition with (D) parts into (D - 1) real-valued coordinates. These transformed coordinates can then be used in standard regression models because they are no longer constrained to sum to a fixed total. In this analysis, the ILR-transformed variables represent balances between groups of compositional parts rather than isolated changes in individual percentages.

Exact zero values cannot be used directly in log-ratio transformations because the logarithm of zero is undefined. Therefore, before applying the ILR transformation, zero percentages were replaced with a very small positive value. This allowed the transformation to be computed while minimally altering the original data structure.

In this study, the ILR transformation was applied to demographic variables describing the composition of urban Small Areas. The transformed ILR coordinates were then used as covariates in the statistical model instead of the original percentage variables. This approach reduces collinearity between demographic percentages and ensures that the regression model respects the relative nature of compositional data.

```{r ilr-transformation, message=FALSE, warning=FALSE}
library(tidyverse)
library(compositions)

# Demographic compositional variables
cols <- c(
  "pct_0_12",
  "pct_13_21",
  "pct_22_40",
  "pct_41_70",
  "pct_71plus",
  "pct_0car",
  "pct_1car",
  "pct_2plus",
  "pct_none_primary",
  "pct_secondary",
  "pct_tertiary",
  "pct_still_in_edu"
)

# Check that all required variables exist in the urban dataset
missing_cols <- setdiff(cols, names(df_urban))

if (length(missing_cols) > 0) {
  stop(
    "These ILR variables are missing from the new dataset:\n",
    paste(missing_cols, collapse = ", ")
  )
}

# Keep missing values as missing
df_urban <- df_urban %>%
  mutate(
    across(
      all_of(cols),
      ~ ifelse(is.na(.x), NA_real_, .x)
    )
  )

# Convert selected percentage variables into a matrix
matrix_urban <- df_urban %>%
  select(all_of(cols)) %>%
  as.matrix()

# Replace exact zeros with a very small positive value
# This is necessary because log-ratio transformations cannot handle zeros
matrix_urban[matrix_urban == 0] <- 0.001

# Apply ILR transformation
variables_ilr <- ilr(acomp(matrix_urban))

# Convert ILR coordinates to a dataframe
variables_ilr_df <- as.data.frame(variables_ilr)

# Name ILR variables
# A composition with D parts produces D - 1 ILR coordinates
names(variables_ilr_df) <- paste0("ilr_", cols[-length(cols)])

# Create the final urban modelling dataset
# Original percentage variables are removed and replaced by ILR coordinates
df_urban_model <- df_urban %>%
  select(-all_of(cols)) %>%
  bind_cols(variables_ilr_df)
```

The resulting dataset, `df_urban_model`, contains the original outcome and accessibility variables, but replaces the raw demographic percentage variables with ILR-transformed coordinates. These ILR variables are then used in the Bayesian regression model to adjust for differences in demographic structure across urban Small Areas.


## 3.4 Statistical Model: Bayesian Beta Regression

The main statistical model was used to estimate the association between school place accessibility and active school travel in urban Small Areas. The outcome variable was the percentage of pupils travelling to school by active modes, defined as walking or cycling. Since this outcome is a proportion bounded between 0 and 1 after rescaling, a beta regression model was used.

Beta regression is appropriate for this analysis because the response variable represents a continuous proportion. Standard linear regression would not be ideal because it could generate predicted values below 0 or above 1 and assumes constant variance across the outcome range. In contrast, the beta distribution is naturally defined on the open interval ((0, 1)), making it suitable for modelling proportional outcomes.

Before fitting the model, the active travel percentage was converted from a percentage to a proportion. Since the beta distribution cannot handle exact values of 0 or 1, the transformed response was adjusted slightly so that all observations lay strictly between 0 and 1.

```{r beta-response-preparation, message=FALSE, warning=FALSE}
df_urban_model <- df_urban_model %>%
  mutate(
    perc_active_prop = perc_active / 100,
    perc_active_beta = (perc_active_prop * (nrow(.) - 1) + 0.5) / nrow(.)
  )

range(df_urban_model$perc_active_beta, na.rm = TRUE)
```

The model was fitted only to urban Small Areas, as these were the main population of interest for the analysis. The demographic covariates were included using standardised ILR-transformed variables. These ILR variables adjust for differences in age structure, car ownership, and educational composition while avoiding the collinearity problems associated with raw percentages.

The school accessibility variables were included using natural cubic spline terms. This allowed the relationship between school accessibility and active travel to be non-linear. This is important because the effect of school place accessibility may not be constant across its full range. For example, increases in accessibility may be strongly associated with active travel at lower values, but the association may flatten at higher values. Natural cubic splines allow this type of flexible relationship while avoiding the assumption of a purely linear effect.

Two spline terms were included:

[
ns(\text{school place density per 1,000 children aged 3--18}, df = 3)
]

and

[
ns(\text{school places per km}^2, df = 3)
]

The first term represents school place availability relative to the local school-age population. The second term represents spatial density of predicted school places per square kilometre. Including both terms allows the model to distinguish between accessibility relative to population demand and the spatial concentration of school places.

The fitted model can be written as:

[
y_i \sim \text{Beta}(\mu_i, \phi)
]

[
\text{logit}(\mu_i) =
\alpha +
\sum_{j=1}^{p} \beta_j Z_{ij}
+
f_1(K_i)
+
f_2(D_i)
]

where (y_i) is the active travel proportion in urban Small Area (i), (\mu_i) is the expected active travel proportion, and (\phi) is the beta precision parameter. The terms (Z_{ij}) represent the standardised ILR demographic covariates. The function (f_1(K_i)) represents the natural spline for school place density per 1,000 children aged 3–18, and (f_2(D_i)) represents the natural spline for school places per square kilometre.

```{r beta-model-formula, message=FALSE, warning=FALSE}
library(brms)
library(splines)

# Linear ILR terms
linear_predictors <- z_ilr_vars

# Natural cubic spline terms for kriging-derived accessibility variables
spline_terms <- c(
  "ns(krig_value_per_1000_3_18, df = 3)",
  "ns(places_density_km2, df = 3)"
)

formula_beta <- as.formula(
  paste(
    "perc_active_beta ~",
    paste(c(linear_predictors, spline_terms), collapse = " + ")
  )
)

formula_beta
```

Weakly informative priors were used for the model parameters. A normal prior with mean 0 and standard deviation 1 was assigned to the regression coefficients. This prior regularises the model by discouraging unrealistically large effects while still allowing the data to influence the posterior estimates. The intercept was assigned a wider normal prior with mean 0 and standard deviation 2.5 on the logit scale. The beta precision parameter, (\phi), was assigned an exponential prior to ensure that it remains positive.

```{r beta-model-priors, message=FALSE, warning=FALSE}
priors <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 2.5), class = "Intercept"),
  prior(exponential(1), class = "phi")
)

priors
```

The Bayesian beta regression model was fitted using the `brms` package. The model was run with four Markov chains, 4,000 iterations per chain, and 1,000 warm-up iterations. The `adapt_delta` parameter was set to 0.97 and the maximum tree depth was set to 12 to improve Hamiltonian Monte Carlo sampling and reduce the risk of divergent transitions.

```{r fit-beta-regression-model, message=FALSE, warning=FALSE}
model_urban_krig <- brm(
  formula = formula_beta,
  data = df_urban_model,
  family = Beta(link = "logit"),
  prior = priors,
  chains = 4,
  iter = 4000,
  warmup = 1000,
  seed = 123,
  cores = 4,
  control = list(
    adapt_delta = 0.97,
    max_treedepth = 12
  )
)
```

The logit link was used so that the expected active travel proportion remains between 0 and 1. Coefficients from the model are therefore estimated on the log-odds scale. For this reason, the main interpretation of the model was based on predicted values and marginal effect plots on the response scale rather than on the raw coefficient values alone.

Model convergence was assessed using standard Bayesian diagnostics. These included (\hat{R}), effective sample size, trace plots, divergent transitions, and posterior predictive checks. Values of (\hat{R}) close to 1 indicate that the chains mixed well. Large effective sample sizes indicate that the posterior summaries are reliable. Posterior predictive checks were used to assess whether the model could reproduce the observed distribution of active travel.

```{r beta-model-summary, message=FALSE, warning=FALSE}
summary(model_urban_krig)
```

```{r beta-model-traceplots, fig.cap="Trace plots for the Bayesian beta regression model. Good mixing across chains indicates that the posterior distribution was explored adequately.", message=FALSE, warning=FALSE}
plot(model_urban_krig)
```

```{r beta-model-posterior-predictive-check, fig.cap="Posterior predictive check for the Bayesian beta regression model. The observed active travel distribution is compared with replicated datasets simulated from the fitted model.", message=FALSE, warning=FALSE}
pp_check(model_urban_krig, ndraws = 100)
```

```{r beta-model-divergences, message=FALSE, warning=FALSE}
nuts_params(model_urban_krig) %>%
  filter(Parameter == "divergent__") %>%
  summarise(n_divergences = sum(Value))
```

The estimated relationship between school place accessibility and active travel was visualised using conditional effect plots. These plots show the expected active travel percentage across the range of each accessibility variable while holding the other covariates constant. They are more interpretable than the raw spline coefficients because spline coefficients do not represent simple one-unit changes.

```{r beta-conditional-effect-krig, fig.cap="Estimated relationship between school place density per 1,000 children aged 3–18 and active school travel in urban Small Areas. The relationship is modelled using a natural cubic spline.", message=FALSE, warning=FALSE}
conditional_effects(
  model_urban_krig,
  effects = "krig_value_per_1000_3_18"
) %>%
  plot(points = TRUE)
```

```{r beta-conditional-effect-density, fig.cap="Estimated relationship between predicted school places per square kilometre and active school travel in urban Small Areas. The relationship is modelled using a natural cubic spline.", message=FALSE, warning=FALSE}
conditional_effects(
  model_urban_krig,
  effects = "places_density_km2"
) %>%
  plot(points = TRUE)
```

Bayesian (R^2) was used as an additional summary of model fit. This gives an approximate measure of the proportion of variation in active travel explained by the model.

```{r beta-model-bayes-r2, message=FALSE, warning=FALSE}
bayes_R2(model_urban_krig)
```

Finally, the fitted model object was saved for later use in the results section, including posterior summaries, model diagnostics, and marginal effect plots.

```{r save-beta-model, message=FALSE, warning=FALSE}
saveRDS(model_urban_krig, "outputs/model_urban_krig.rds")
```

Overall, this model estimates the adjusted association between school place accessibility and active school travel in urban Small Areas. The model accounts for demographic composition through standardised ILR-transformed covariates and allows non-linear associations for the two school accessibility variables using natural cubic splines. The results are interpreted using posterior credible intervals and predicted values on the response scale.


## 3.5 Normalisation by School-Age Population

After estimating the raw kriging-derived school place intensity, the measure was normalised by the local school-age population. This step was necessary because the raw kriging value represents predicted school place availability, but it does not account for the number of children living in each Small Area. Areas with higher predicted school place intensity may not necessarily have better accessibility if they also contain a large number of school-age children.

To account for local demand, the raw kriged school place value was divided by the population aged 3 to 18 years in each Small Area. This age range was selected because it approximates the population potentially requiring access to primary, post-primary, and related school services. The resulting measure was then multiplied by 1,000 to make the variable easier to interpret.

The final normalised school place accessibility measure was calculated as:

[
\text{School place density per 1,000 children}_{i}
==================================================

1000 \times
\frac{
\text{Raw kriged school place value}*{i}
}{
\text{Population aged 3--18}*{i}
}
]

where (i) represents a Small Area.

Higher values indicate that an area has greater predicted school place availability relative to the number of children aged 3–18 living there. Lower values indicate that predicted school place availability is lower relative to the local school-age population. Therefore, this variable can be interpreted as a relative school accessibility measure rather than a simple measure of school concentration.

This normalisation was important for the research question because active school travel is likely to depend not only on the presence of schools, but also on the relationship between school availability and the local population of children. Without this adjustment, areas with high absolute school place intensity could be overestimated as highly accessible, even if they also have a large school-age population.

```{r normalisation-school-age-population, message=FALSE, warning=FALSE} id="qfeg4z"
final_sf <- final_sf %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    
    # Raw kriging value kept for reference
    krig_value_raw = krig_value,
    
    # Normalise by the population aged 3 to 18
    krig_value_per_child_3_18 = safe_divide(
      krig_value_raw,
      pop_3_18
    ),
    
    # Express as predicted school place intensity per 1,000 children aged 3 to 18
    krig_value_per_1000_3_18 = 1000 * krig_value_per_child_3_18,
    
    # Use the normalised variable as the main kriging exposure
    krig_value = krig_value_per_1000_3_18,
    
    # Additional spatial density variable
    places_density_km2 = safe_divide(
      krig_value_raw,
      area_km2
    )
  )
```

The variable `krig_value_raw` was retained as the unnormalised kriging prediction. The variable `krig_value_per_child_3_18` represents the predicted school place intensity per child aged 3–18. The variable `krig_value_per_1000_3_18` expresses the same measure per 1,000 children, making it more interpretable. In the final modelling dataset, `krig_value` was set equal to `krig_value_per_1000_3_18`, meaning that the main school accessibility variable used in the statistical model was the normalised measure.

```{r normalisation-summary, message=FALSE, warning=FALSE} id="hglpml"
final_df %>%
  summarise(
    n = n(),
    min_population_3_18 = min(pop_3_18, na.rm = TRUE),
    median_population_3_18 = median(pop_3_18, na.rm = TRUE),
    max_population_3_18 = max(pop_3_18, na.rm = TRUE),
    min_school_density = min(krig_value_per_1000_3_18, na.rm = TRUE),
    median_school_density = median(krig_value_per_1000_3_18, na.rm = TRUE),
    max_school_density = max(krig_value_per_1000_3_18, na.rm = TRUE)
  )
```

For interpretation, this normalised variable should be understood as a relative measure of school place availability. It does not represent the exact number of school places physically located within each Small Area. Instead, it represents the kriging-derived estimate of school place intensity relative to the number of children aged 3–18 living in that area.


# 4 Results

## 4.1 Spatial Distribution of Schools

## 4.2 Walking and Cycling Accessibility: travel time

## 4.3 Kriging vs Travel Time Comparison

## 4.4 Model Results — Urban Areas

## 4.5 Effect of Walking Accessibility on Active Travel

# 5 Discussion

## 5.1 Interpretation of Results

## 5.2 Limitations

## 5.3 Comparison with the article

# 6 Conclusions

# 7 References