---
title: "Distance, Mobility, and Mode Choice : A Spatial Analysis of School Travel Behaviour in Ireland"
authors:
  - name: Sananthaa Jagadesan Sethamaraikannan
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland. 
    email: s.jagadesansenthamaraikannan1@universityofgalway.ie
  - name: Judith Beltran Lopez
    department: Department of Mathematics & Statistics
    affiliation: University of Galway
    location: Galway,Ireland
    email: j.beltranlopez1@universityofgalway.ie
abstract: |
  Active travel to school is shaped by the spatial relationship between where children live, where schools are located, and how easily schools can be reached. This study investigates whether school-place accessibility is associated with differences in the proportion of children walking or cycling to school across Ireland. Data from Census 2022, the Department of Education, and OpenStreetMap were combined to construct school accessibility measures for Census Small Areas. Two complementary accessibility measures were developed: a kriging-based measure of school-place density, based on straight-line distance, and a routing-based walking accessibility measure, based on the number of school places reachable on foot within 30 minutes along the road and path network. Bayesian beta regression was used to model the association between school accessibility and the proportion of children travelling to school by active modes, adjusting for demographic composition using isometric log-ratio transformed covariates. The two measures agreed on average but diverged in many individual areas. These divergences consistently reflected severance, where physical barriers such as motorways, rivers, and water cut an area off from schools that are close in straight-line terms but not reachable on foot. The walking accessibility measure was much more strongly associated with active travel than the straight-line kriging measure, suggesting that whether schools can actually be reached on foot matters more for active school travel than how close they are in straight-line distance.
keywords:
  - active travel
  - school accessibility
  - school-place density
  - kriging
  - travel-time accessibility
  - Census Small Areas
  - Bayesian beta regression
  - Ireland
bibliography: references.bib
biblio-style: unsrt
header-includes:
  - \providecommand{\pandocbounded}[1]{#1}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{float}
  - \floatplacement{figure}{!ht}
  - \setlength{\textfloatsep}{10pt plus 2pt minus 4pt}
  - \setlength{\floatsep}{10pt plus 2pt minus 4pt}
  - \setlength{\intextsep}{10pt plus 2pt minus 4pt}
  - \usepackage{parskip}
  - \setlength{\parskip}{4pt}
  - \usepackage{titlesec}
  - \titlespacing{\section}{0pt}{10pt}{6pt}
  - \titlespacing{\subsection}{0pt}{8pt}{4pt}
  - \titlespacing{\subsubsection}{0pt}{6pt}{3pt}
output: rticles::arxiv_article
---

# Introduction
Active travel to school, particularly walking and cycling, is an important part of everyday mobility for children and young people. It is shaped not only by individual or household characteristics, but also by the spatial organisation of schools, neighbourhoods, and transport networks. Children are more likely to walk or cycle when schools are located within a feasible distance and can be reached through safe and connected routes. For this reason, school accessibility is an important factor in understanding differences in travel behaviour to education.

In Ireland, access to nearby schools is not evenly distributed across space. Some areas have a high concentration of school places within short walking or cycling distances, while others have fewer nearby options. This has been described as a form of “postcode lottery”, where opportunities for active school travel may depend partly on where children live. However, this argument requires spatial evidence: it is necessary to measure school accessibility at a local scale and examine whether areas with better access to school places also have higher active travel rates.

This study investigates the relationship between school-place accessibility and active school travel across Ireland. The analysis combines Census 2022 Small Area data, Department of Education school enrolment and location data, and OpenStreetMap network data. Two complementary measures of school accessibility are developed. First, a kriging-based measure estimates the spatial distribution of school-place density across Census Small Areas using straight-line distance. Second, a routing-based travel-time measure estimates how many school places can be reached on foot within a travel-time threshold using the road and path network.

The main outcome is the proportion of pupils travelling to school by active modes, defined as walking or cycling. Bayesian beta regression is used to model the association between school accessibility and active travel in urban Census Small Areas, while adjusting for demographic composition, car availability, and educational attainment. This allows the analysis to examine whether differences in school accessibility are associated with differences in active school travel after accounting for local population characteristics.The objectives of this study are therefore to: construct local measures of school-place accessibility across Ireland; compare kriging-based and travel-time-based accessibility measures; estimate the association between school accessibility and active school travel; and assess whether spatial inequalities in school-place accessibility may contribute to unequal opportunities for walking and cycling to school.


## Background
School travel is not only a question of individual choice. It is also shaped by distance, school location, educational provision, and the structure of the local transport network. Walking and cycling to school are widely promoted because they can support healthier daily routines, reduce traffic congestion, and contribute to lower transport-related emissions. However, the opportunity to travel actively depends on whether schools are located within a feasible distance and whether they can be reached through suitable routes.

In Ireland, low levels of active travel to school are often discussed in terms of behaviour change, with policy attention directed towards encouraging pupils and families to walk or cycle. While behavioural factors are important, this explanation can overlook the spatial conditions that make active travel possible. If children do not have access to nearby schools, or if available schools are not reachable within reasonable walking or cycling times, then active travel may be difficult regardless of individual preferences.

Recent research has highlighted the importance of educational provision in shaping school travel patterns. Mancini argues that inequalities in school provision and school access form part of the wider Irish educational landscape [@mancini2024children]. Related work on adolescent active transport in Ireland suggests that active travel does not simply follow population size or density, but is strongly linked to the presence and type of local school provision [@mancini2025census]. This argument has also been discussed in a public-facing RTÉ Brainstorm article, which describes teenage active transport as a potential “postcode lottery” shaped by where young people live and what schools are available locally [@mancini2024rte].

This study builds on that research context by measuring school accessibility at a finer spatial scale. Instead of analysing only settlement-level differences, it uses Census Small Areas to examine local variation in school-place accessibility and active school travel. By combining school enrolment and location data, Census 2022 travel data, and OpenStreetMap network data, the study assesses whether areas with greater access to school places also have higher proportions of pupils walking or cycling to school.


## The Postcode Lottery Argument from the Article
The idea of a “postcode lottery” provides an important motivation for this study. In the context of school travel, the term suggests that opportunities for walking or cycling to school may depend partly on where a child lives. If school places are unevenly distributed, then children in some neighbourhoods may have nearby schools that are reachable by active modes, while children in other areas may need to travel longer distances or rely on motorised transport.

Mancini’s work argues that active travel to school should not be understood only as a matter of individual behaviour or family preference. Instead, it is also shaped by the geography of school provision. The RTÉ Brainstorm article summarising this research states that disparities between areas are not simply explained by population size, but are linked to differences in educational provision [@mancini2024rte]. Low levels of walking or cycling are not simply the result of unwillingness to travel actively. In some places, active travel may be less feasible because suitable school places are not located nearby or are not reachable through the local transport network.

This argument is also connected to wider research on inequalities in the Irish educational landscape. Mancini’s study of unequal school provision examines how the distribution and type of schools can create different educational opportunities across communities [@mancini2024children]. Related research on adolescent active transport in Ireland links school provision with unequal access to walking and cycling opportunities, showing that the spatial organisation of schools is relevant to sustainability, mobility, and education policy [@mancini2025census].

In this study, the “postcode lottery” argument is treated as an empirical question. Rather than assuming that school access directly determines active travel, the analysis tests whether areas with greater school-place accessibility also show higher proportions of pupils walking or cycling to school. This is done by constructing two complementary accessibility measures: a kriging-based measure of school-place density and a routing-based travel-time measure of reachable school places. By analysing these measures at Census Small Area level, the study examines whether spatial inequalities in school provision are associated with unequal opportunities for active school travel in Ireland.


## Research Objectives
Working with official statistics presents important methodological challenges for studying school travel behaviour at a detailed spatial scale. Census and administrative datasets are usually released in aggregated form, meaning that individual residential locations, school destinations, and travel routes are not directly observed. This study therefore combines spatial modelling, accessibility analysis, and Census Small Area data to examine how school-place accessibility relates to active school travel across Ireland.

The first objective is to map and analyse the spatial distribution of schools across Ireland, using school location and enrolment data to describe where school places are concentrated.The second objective is to develop local measures of school accessibility. This includes a kriging-based measure of school-place density and a routing-based travel-time measure of the number of school places reachable on foot.The third objective is to describe school travel behaviour using Census 2022 data, focusing on the prevalence of active travel modes, defined as walking and cycling, across Census Small Areas.The fourth objective is to examine the relationship between school accessibility and active school travel in urban Census Small Areas, while adjusting for demographic composition, car availability, and educational attainment.Together, these objectives aim to assess whether spatial differences in school-place accessibility are associated with unequal opportunities for walking and cycling to school in Ireland.


# Data

## School Data
School data were obtained from the Department of Education and Youth's 2024/25 school datasets. Primary school records were taken from the mainstream schools enrolment dataset [@departmenteducation_primary_2024_25], and post-primary school records from the corresponding post-primary dataset, which included Eircodes [@departmenteducation_postprimary_2024_25]. Both datasets provide school-level information, including location coordinates and enrolment figures. The 2024/25 datasets were used because they provide recent enrolment information together with latitude and longitude coordinates for each school, which allowed schools to be represented directly as spatial point data without requiring additional geocoding.

The two datasets were combined into a single set of school points, giving 3,081 primary schools and 723 post-primary schools, a combined total of 3,804 schools used in the analysis. Each record represents an individual school and includes its location and enrolment. Enrolment was used as a proxy for the number of school places available at each location. Although enrolment does not measure spare capacity directly, it provides a consistent measure of the relative size of schools and the spatial distribution of school places across the country.

The key variables extracted were school latitude, school longitude, and enrolment. Records without valid coordinates, or without a valid positive enrolment value, were removed because they could not be used in the spatial analysis. After this step, enrolment values ranged from one to just over one thousand pupils per school, with no zero values remaining.
All school coordinates were initially provided in latitude and longitude using the WGS84 coordinate reference system. The school points were then transformed to the Irish Transverse Mercator projection, EPSG:2157. This projected coordinate system was used because distances, areas, and spatial operations are measured in metres, making it appropriate for the kriging analysis and for linking school locations to Census Small Areas.

Where multiple school records shared identical coordinates, their enrolments were summed before analysis. This affected co-located schools, such as junior and senior schools on the same campus, which then contributed their combined enrolment to the school-place measure. Summing duplicate coordinates in this way avoided duplicated spatial locations while preserving the total number of pupils represented at each site.

The cleaned school point dataset was then used in two complementary parts of the accessibility analysis. First, the school locations and enrolment values were used as the observed input for the kriging-based school-place density measure, producing an area-level estimate of school-place availability across Census Small Areas. Second, the same school locations were used as destination points in the routing-based travel-time accessibility analysis. In that analysis, school enrolment was used to weight destinations, so that accessibility was measured as the number of school places reachable on foot within a travel-time threshold, rather than simply the number of schools reachable.
The school dataset therefore contributed both a spatial density measure and a network-based travel-time measure of school accessibility. The kriging-based measure captures the broader spatial distribution of school places, while the routing-based measure captures how many school places can be reached through the actual road and path network within a given walking time.

## Census Data
Census data were obtained from the Central Statistics Office (CSO) Census 2022 Small Area Population Statistics (SAPS) dataset [@cso_saps_2022]. The analysis was carried out at the Census Small Area (CSA) level. The CSA is the smallest geography at which Census data are released, which makes it well suited to studying school travel, because walking and cycling to school are short, local journeys, and a fine spatial unit captures local variation that larger settlement-level geographies would average out. CSAs can also be directly linked to the school accessibility measures constructed from school locations, kriging predictions, and routing-based travel-time calculations.

The main outcome variable was the proportion of people travelling to education by active modes, defined as walking or cycling. It was calculated from the Census 2022 travel-to-education variables, which record how those attending school or college usually travel, excluding records with no stated mode of travel or for whom the question did not apply. This produced an active-travel percentage for each CSA, which was linked to the school accessibility measures and used as the response variable in the statistical model.

The SAPS dataset was also used to derive demographic and population covariates, including total population, population aged 3–18 years, age composition, household car availability, educational attainment, and urban–rural classification. The population aged 3–18 years was used to express both accessibility measures relative to the number of school-age children living in each CSA. The CSO classifies each CSA into one of six urban–rural categories. For this analysis, cities, satellite urban towns, and independent urban towns were treated as urban, while the three rural categories (rural areas with high, moderate, and low urban influence) were treated as rural. The main statistical model was restricted to urban CSAs, where population density, school density, and street connectivity make walking and cycling to school more feasible.

## Study Area
The study area was the Republic of Ireland. The analysis was conducted at the Census Small Area level, using Census Small Area boundaries from the 2022 national statistical boundary dataset [@tailte_smallareas_2022]. Census Small Areas were selected as the main spatial unit because they provide a fine geographical scale for linking population characteristics, school accessibility, and travel behaviour. This scale is more appropriate for studying school travel than larger settlement-level geographies, because walking and cycling to school are highly local behaviours.

```{r study-area-csa, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, out.width="100%", fig.cap="Census Small Areas (CSAs) across Ireland (left) and zoomed in on Galway (right), shaded by area. CSAs are small in densely populated urban areas and large in sparsely populated rural areas."}

library(tidyverse)
library(sf)
library(scales)
library(patchwork)

obj <- readRDS("../../outputs/combined_kriging_objects.rds")
final_sf <- obj$final_sf
ireland_boundary <- obj$dat_irl$region_sf %>% st_transform(st_crs(final_sf)) %>% st_make_valid()

counties_sf <- st_read("../../data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE) %>%
  st_transform(st_crs(final_sf)) %>% st_make_valid()
galway_boundary <- counties_sf %>%
  filter(stringr::str_detect(NAME_1, regex("Galway", ignore_case = TRUE)) |
         stringr::str_detect(NAME_2, regex("Galway", ignore_case = TRUE))) %>%
  st_union() %>% st_as_sf() %>% mutate(county = "Galway")

# add area (km2) for colouring, capped so a few huge rural CSAs don't dominate
csa_area <- final_sf %>%
  mutate(area_km2 = as.numeric(st_area(.)) / 1e6,
         area_cap = pmin(area_km2, quantile(area_km2, 0.95, na.rm = TRUE)))

galway_area <- csa_area[galway_boundary, ]

p_nat <- ggplot() +
  geom_sf(data = csa_area, aes(fill = area_cap), colour = NA) +
  geom_sf(data = ireland_boundary, fill = NA, colour = "grey30", linewidth = 0.3) +
  scale_fill_viridis_c(name = "CSA area (km2)", trans = "sqrt",
                       labels = label_number(big.mark = ",")) +
  labs(title = "CSAs across Ireland") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")

p_zoom <- ggplot() +
  geom_sf(data = galway_area, aes(fill = area_cap), colour = "grey40", linewidth = 0.1) +
  geom_sf(data = galway_boundary, fill = NA, colour = "black", linewidth = 0.4) +
  scale_fill_viridis_c(name = "CSA area (km2)", trans = "sqrt",
                       labels = label_number(big.mark = ",")) +
  labs(title = " CSAs in Galway") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")

p_nat + p_zoom + plot_layout(guides = "collect")
```

Figure 1 shows the study area and the Census Small Area (CSA) unit. The left panel shows all CSAs across Ireland, and the right panel zooms in on Galway so the individual areas can be seen. Each CSA is shaded by its size in square kilometres. CSAs are designed to contain roughly the same number of people, so their geographic size depends on how densely the area is populated: in cities and towns, where many people live close together, CSAs are small (darker colours), while in the countryside, where people are more spread out, CSAs are much larger (lighter colours). This is why cities appear as small dark clusters and rural areas as large pale shapes.

Each Census Small Area was treated as a neighbourhood-level unit of analysis. Census 2022 demographic and travel variables were linked to these boundaries, while school accessibility measures were derived from school locations and then summarised at the Census Small Area level. For the kriging-based measure, school-place density predictions were aggregated to Census Small Areas. For the routing-based travel-time measure, Census Small Areas were used as the origins from which reachable school places were calculated using the road and path network.

All spatial layers were transformed to the Irish Transverse Mercator coordinate reference system, EPSG:2157, before distance, area, and accessibility calculations were carried out. Using a projected coordinate system ensured that spatial measurements were calculated in metres and were consistent across the analysis.The main statistical modelling was restricted to urban Census Small Areas. This restriction was used because the research question focuses on active school travel, and walking and cycling are more feasible in urban settings where population density, school density, and street-network connectivity are generally higher. Rural Census Small Areas were retained for descriptive spatial comparisons, but they were not included in the main Bayesian beta regression model.

## Constructing the Final Dataset
The final dataset was assembled by bringing the school, census, and network data together at the Census Small Area (CSA) level. School locations from the primary and post-primary files were combined into a single set of enrolment-weighted points, which served as the input for both accessibility measures. For the kriging measure, school-place density was predicted at a 3 × 3 grid of points within each CSA and averaged to give one value per area. For the walking measure, the same CSA grid points were used as origins, and travel times to the school locations were computed over the OpenStreetMap network, counting the enrolment-weighted school places reachable on foot within 30 minutes and averaging to one value per CSA. Both measures were then normalised by the CSA population aged 3–18 to express accessibility per 100 children.

These accessibility measures were joined to the Census active-travel outcome and the demographic covariates, all at CSA level, to form a single dataset. Some CSAs were then removed so that the modelling sample was reliable: areas with missing required values, areas with a total population above 900, and areas with fewer than five children aged 3–18 were excluded, the last two to avoid unstable estimates from unusually large areas or very small denominators. The main statistical model was then restricted to the urban CSAs within this cleaned dataset, giving the final modelling sample used in the analysis.

# Methods

## School Place Density: Kriging
A kriging-based school-place density measure was developed to estimate spatial variation in school-place availability across Census Small Areas. Kriging is a geostatistical interpolation method that uses the spatial correlation structure of observed point data to predict values at unsampled locations. In this study, the observed point data were school locations, and the variable interpolated was school enrolment, used as a proxy for school-place availability. The general kriging workflow, including empirical variograms, fitted variogram models, and spatial prediction, follows standard geostatistical practice and is described in Moraga’s spatial statistics text [@moraga2023spatial; @moraga2023kriging]. The kriging implementation was carried out using the `gstat` package in R [@pebesma2004gstat].

School records were first converted to spatial points using their latitude and longitude coordinates and were then transformed to the Irish Transverse Mercator coordinate reference system, EPSG:2157. This ensured that distance-based calculations were carried out in metres. Records without valid coordinates or valid enrolment values were removed because they could not contribute to the spatial interpolation. Where multiple schools shared exactly the same coordinates, their enrolments were summed before analysis. This ensured that co-located schools, such as junior and senior schools on the same campus, contributed their combined enrolment to the school-place measure. After aggregation, the total enrolment at each unique school location was log-transformed before fitting the kriging model.

Ordinary kriging was used to predict school-place availability at unsampled locations. The spatial dependence in log-enrolment was first assessed using an empirical variogram. The empirical variogram shows how dissimilarity in log-enrolment changes as the distance between pairs of school locations increases. A fitted variogram model was then used to describe this spatial dependence formally and to provide the covariance structure required for kriging prediction. In the final analysis, a Gaussian variogram model was fitted to the empirical variogram. The variogram was estimated using a lag width of 5 km and a maximum cutoff distance of 200 km. Kriging predictions were made using up to 150 nearby observed school locations for each prediction point.

Instead of predicting only at the centroid of each Census Small Area, each Census Small Area polygon was divided into a 3 × 3 internal grid. Grid cells were clipped to the Census Small Area boundary, and one internal point was created within each valid clipped grid cell using a point-on-surface approach. Kriging predictions were then obtained at these internal prediction points and averaged to produce one raw kriging-derived estimate for each Census Small Area. This approach reduced dependence on a single centroid and better represented irregularly shaped areas.Because the kriging model was fitted on the log-enrolment scale, predictions were transformed back to the original enrolment scale before aggregation. A bias-corrected log-normal back-transformation was used:

$$
\widehat{E}_i =
\exp \left(
\hat{z}_i + \frac{1}{2}\hat{\sigma}^{2}_i
\right)
$$

where $\hat{z}_i$ is the kriging prediction on the log scale at prediction point (i), and $\hat{\sigma}^{2}_i$ is the kriging prediction variance at that point. This correction was used because simply exponentiating the predicted log value would underestimate the expected value on the original enrolment scale.

The back-transformed predictions were averaged across the internal prediction points within each Census Small Area to produce one raw school-place estimate per area. The corresponding kriging prediction variance was also averaged across the prediction points to provide an area-level measure of interpolation uncertainty.

The raw kriging estimate represents predicted school-place availability, but it does not account for the number of children living in each area. To account for local demand, the raw kriged value was divided by the Census 2022 population aged 3–18 years in each Census Small Area. The final normalised school-place density measure was expressed per 100 children aged 3–18:

$$
SchoolDensity_i =
100 \times
\frac{RawKrigedValue_i}{PopulationAged3to18_i}
$$

where $RawKrigedValue_i$ is the raw kriging-derived school-place estimate for Census Small Area (i), and (PopulationAged3to18_i) is the number of residents aged 3–18 in that area. Higher values indicate greater predicted school-place availability relative to the local school-age population. Lower values indicate lower predicted availability relative to potential local demand. This normalised measure was used as the main kriging-based school accessibility exposure in the statistical model. The unnormalised kriging estimate was retained as a reference measure of absolute predicted school-place availability, and school-place density per square kilometre was also calculated as a secondary spatial density measure.

### Kriging diagnostic figures
The following figures were used to check the spatial input data, the fitted variogram, and the resulting kriging-derived school-place density surface. These figures are included as diagnostics and descriptive outputs rather than as part of the model-fitting code.

```{r kriging-plot-setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(scales)
library(stringr)
library(gstat)

obj <- readRDS("../../outputs/combined_kriging_objects.rds")

dat_irl <- obj$dat_irl
variogram_emp <- obj$variogram
variogram_model <- obj$variogram_model
final_sf <- obj$final_sf
final_df <- obj$final_df
final_df_urban <- obj$final_df_urban

final_sf_plot <- final_sf %>%
  st_make_valid() %>%
  filter(
    is.finite(krig_value_raw),
    is.finite(krig_value_per_100_3_18),
    is.finite(pop_3_18)
  ) %>%
  mutate(
    krig_raw_cap = pmin(
      krig_value_raw,
      quantile(krig_value_raw, 0.99, na.rm = TRUE)
    ),
    krig_per100_cap = pmin(
      krig_value_per_100_3_18,
      quantile(krig_value_per_100_3_18, 0.99, na.rm = TRUE)
    )
  )

schools_plot <- dat_irl$schools_agg_sf %>%
  st_transform(st_crs(final_sf_plot)) %>%
  mutate(
    enrolment_plot = case_when(
      "enrolment_total" %in% names(.) ~ enrolment_total,
      "log_enrol" %in% names(.) ~ exp(log_enrol),
      TRUE ~ NA_real_
    )
  )

ireland_boundary <- dat_irl$region_sf %>%
  st_transform(st_crs(final_sf_plot)) %>%
  st_make_valid()

counties_shp_path <- "../../data/gadm41_IRL_shp/gadm41_IRL_2.shp"

counties_sf <- st_read(counties_shp_path, quiet = TRUE) %>%
  st_transform(st_crs(final_sf_plot)) %>%
  st_make_valid()

galway_boundary <- counties_sf %>%
  filter(
    str_detect(NAME_1, regex("Galway", ignore_case = TRUE)) |
      str_detect(NAME_2, regex("Galway", ignore_case = TRUE))
  ) %>%
  st_union() %>%
  st_as_sf() %>%
  mutate(county = "Galway")

galway_sf_plot <- final_sf_plot[galway_boundary, ]
schools_galway <- schools_plot[galway_boundary, ]
```

```{r school-locations-kriging-input, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of school locations used as input for the kriging analysis. Points represent aggregated school locations, with point size proportional to total enrolment at each location."}
ggplot() +
  geom_sf(
    data = ireland_boundary,
    fill = "grey97",
    colour = "grey60",
    linewidth = 0.2
  ) +
  geom_sf(
    data = schools_plot,
    aes(size = enrolment_plot),
    alpha = 0.55
  ) +
  scale_size_continuous(
    name = "Enrolment",
    range = c(0.2, 2.2),
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "School locations used for kriging",
    subtitle = "Aggregated school locations weighted by total enrolment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r kriging-variogram-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Empirical and fitted variogram for log-enrolment. The x-axis shows distance between pairs of school locations, and the y-axis shows semivariance in log-enrolment."}
plot(
  variogram_emp,
  variogram_model,
  main = "Empirical and fitted variogram for log-enrolment"
)
```

```{r normalised-kriging-ireland-map, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Normalised kriged school-place density by Census Small Area. Values are expressed per 100 children aged 3–18. The colour scale is capped at the 99th percentile to make spatial patterns easier to see."}
ggplot() +
  geom_sf(
    data = final_sf_plot,
    aes(fill = krig_per100_cap),
    colour = NA
  ) +
  geom_sf(
    data = ireland_boundary,
    fill = NA,
    colour = "grey40",
    linewidth = 0.15
  ) +
  scale_fill_viridis_c(
    name = "School-place density\nper 100 children",
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "Normalised kriged school-place density",
    subtitle = "Per 100 children aged 3–18; colour capped at 99th percentile"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r kriging-density-distribution, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Distribution of the normalised kriging-based school-place density measure. The x-axis is limited at the 99th percentile to reduce the influence of extreme values in the plot."}
final_df %>%
  filter(is.finite(krig_value_per_100_3_18)) %>%
  ggplot(aes(x = krig_value_per_100_3_18)) +
  geom_histogram(bins = 40) +
  coord_cartesian(
    xlim = c(
      0,
      quantile(final_df$krig_value_per_100_3_18, 0.99, na.rm = TRUE)
    )
  ) +
  labs(
    title = "Distribution of kriged school-place density",
    subtitle = "Per 100 children aged 3–18; x-axis capped at 99th percentile",
    x = "School-place density per 100 children aged 3–18",
    y = "Number of Census Small Areas"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold")
  )
```

```{r galway-kriging-school-overlay, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Example local view of the normalised kriging-based school-place density measure in Galway. School points are overlaid to show how the interpolated surface relates to observed school locations."}
ggplot() +
  geom_sf(
    data = galway_sf_plot,
    aes(fill = krig_per100_cap),
    colour = NA
  ) +
  geom_sf(
    data = galway_boundary,
    fill = NA,
    colour = "black",
    linewidth = 0.4
  ) +
  geom_sf(
    data = schools_galway,
    aes(size = enrolment_plot),
    alpha = 0.7
  ) +
  scale_fill_viridis_c(
    name = "School-place density\nper 100 children",
    labels = label_number(big.mark = ",")
  ) +
  scale_size_continuous(
    name = "School enrolment",
    range = c(0.5, 3),
    labels = label_number(big.mark = ",")
  ) +
  labs(
    title = "Galway kriging values with school locations",
    subtitle = "Normalised kriging per 100 children aged 3–18"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

The national map shows the broad spatial pattern in the kriging-derived school-place density measure, while the Galway map provides a clearer local example of how observed school locations contribute to the interpolated surface. The distribution plot is included because the normalised kriging variable is highly skewed; for this reason, later descriptive plots and maps use percentile capping where appropriate to improve readability.

### Normalisation by School-Age Population
The raw kriging-derived estimate represents predicted school-place availability, but it does not account for the number of children living in each Census Small Area. This adjustment was necessary because an area with a high predicted number of school places may not necessarily have high accessibility if it also has a large school-age population. To account for local demand, the raw kriged school-place estimate was normalised by the Census 2022 population aged 3–18 years in each Census Small Area [@cso_saps_2022].

The population aged 3–18 years was used because it approximates the population potentially requiring access to primary and post-primary education. Although this age range does not correspond exactly to school attendance in every case, it provides a consistent local denominator for comparing school-place availability across Census Small Areas.

The normalised kriging-based accessibility measure was calculated as:

$$
SchoolDensity_i =
100 \times
\frac{RawKrigedValue_i}{PopulationAged3to18_i}
$$

where $RawKrigedValue_i$ is the raw kriging-derived school-place estimate for Census Small Area $i$, and $PopulationAged3to18_i$ is the number of residents aged 3–18 in that area. Multiplying by 100 expresses the measure as predicted school-place availability per 100 children aged 3–18, which is easier to interpret at the Census Small Area scale than a per-child value.

Higher values indicate greater predicted school-place availability relative to the local school-age population. Lower values indicate lower predicted availability relative to potential local demand. This means that the normalised measure should be interpreted as a relative accessibility measure, rather than as the exact number of school places physically located within each Census Small Area.

The unnormalised kriging value was retained as a reference measure of absolute predicted school-place availability. An additional spatial density variable, predicted school places per square kilometre, was also calculated to describe the concentration of school places independently of the local school-age population denominator. The normalised measure per 100 children aged 3–18 was used as the main kriging-based accessibility exposure in the Bayesian beta regression model.


## Travel-Time Accessibility

### Rationale and choice of measure
The kriging-based measure described in the previous section estimates how densely school places are distributed around a location, but it does so in straight-line (Euclidean) space and takes no account of the routes a child would actually use. A high kriged density can occur in an area that is in practice difficult to traverse on foot, for example where a river, motorway, or absent crossing separates homes from nearby schools. To capture accessibility as it is experienced along the real network, a second measure was constructed based on travel time rather than straight-line proximity. This measure asks how many school places a child could actually reach from their own area within a feasible walking time, travelling along the existing road and path network.

This is a cumulative-opportunity accessibility measure, one of the standard place-based approaches to measuring accessibility, in which the number of opportunities reachable from an origin within a fixed travel-time threshold is counted [@geurs2004accessibility]. This type of measure is well suited to school travel, because journeys to school are short and local, and the relevant question is not the average distance to all schools but whether a reasonable number of school places fall within the limited time a child would realistically spend walking. By counting reachable places within a plausible time limit, the measure reflects the practical opportunity for active travel in each area rather than straight-line proximity alone.

### The underlying method
The measure was produced using network-based routing, in which shortest-path travel times are computed across a street network from each origin to a set of candidate destinations. For every origin–destination pair, the routing algorithm finds the quickest path through the network for the chosen travel mode, and destinations whose travel time falls within a chosen threshold are then treated as reachable. This is the method underlying the accessibility calculations, and it is distinct from any particular software used to carry it out.

In this study the routing was implemented using the R5 routing engine, accessed in R through the r5r package [@pereira2021r5r; @conway2018r5]. R5 computes travel times over a network derived from OpenStreetMap data and can handle large numbers of origin–destination pairs efficiently, which made it suitable for computing walking accessibility for every Census Small Area in the study.

###  Network and input data
The routing network was built from an OpenStreetMap extract covering Ireland and Northern Ireland, obtained from Geofabrik [@openstreetmap; @geofabrik]. OpenStreetMap provides the road and path geometry, including footpaths and minor links relevant to walking, that R5 uses to construct the routable network. The origins for the accessibility calculation were the Census Small Area prediction points (a 3 × 3 grid of points within each area, as described for the kriging measure), and the destinations were the aggregated school point locations described in the School Data section. Each destination was weighted by its total enrolment, so that a school location contributed its number of school places rather than counting as a single school. This ensured that accessibility reflected the volume of school places reachable from each area, consistent with the enrolment-weighted treatment of schools used throughout the analysis.

###  Travel-time computation and thresholds
Travel times were computed for walking, using R5's default walking speed of 3.6 km/h. Walking accessibility was evaluated at thresholds of 10, 20, and 30 minutes. These thresholds were chosen to span the range of times a child might plausibly spend walking to school, from a short journey up to a realistic upper limit. A step decay function was used, so that all school places within a given threshold were counted equally. For each origin and each threshold, the number of reachable school places was obtained by summing the enrolment-weighted destinations whose network travel time fell within the threshold.

The 30-minute walking measure was carried forward as the primary travel-time exposure in the statistical model. Thirty minutes was treated as the most plausible upper bound for a routine walk to school, long enough to capture the realistic extent of walking journeys while remaining within the time a daily school commute on foot would reasonably take. The shorter 10- and 20-minute thresholds were retained for descriptive comparison but were not used as the main model exposure.

### Normalisation and transformation
As with the kriging measure, the raw count of reachable school places reflects the supply of places without reference to the number of children competing for them. To express accessibility relative to local demand, the count of reachable places for each Census Small Area was divided by its Census 2022 population aged 3–18 and expressed per 100 children:

$$
WalkAccess_i = 100 \times \frac{ReachablePlaces_i}{PopulationAged3to18_i}
$$

where $ReachablePlaces_i$ is the enrolment-weighted number of school places reachable within 30 minutes of walking from Census Small Area $i$, and $PopulationAged3to18_i$ is the number of residents aged 3–18 in that area. This matches the per-100-children scaling used for the kriging-based measure, so that both accessibility measures are expressed on a comparable demand-adjusted basis.
 
The distribution of reachable-place counts was strongly right-skewed, because a small number of dense urban areas could reach many thousands of school places while peripheral areas could reach very few. To prevent these extreme values from dominating the model and to allow the predictor to vary on a multiplicative rather than additive scale, the normalised measure was log-transformed using a base-10 logarithm with an offset of one:

$$
WalkAccessLog_i = \log_{10}\left(WalkAccess_i + 1\right)
$$

The offset of one ensured that areas able to reach no school places within the threshold remained defined after transformation. This log-transformed, demand-adjusted variable was used as the final travel-time accessibility exposure in the Bayesian beta regression model.

## Demographic Variables: ILR Transformation
Several demographic variables derived from Census 2022 Small Area Population Statistics were included to adjust for differences in population structure across Census Small Areas. These variables describe age structure, household car availability, and educational attainment. Because these variables are expressed as percentages within groups, they are compositional: the parts represent shares of a whole and are therefore constrained to sum to a fixed total. Standard regression models can give unstable or misleading results when compositional variables are included directly, because an increase in one component necessarily implies a decrease in at least one other component [@aitchison1986compositional].

Three demographic compositions were constructed. Age composition was grouped into five categories: children aged 0–12, young people aged 13–24, adults aged 25–44, adults aged 45–74, and adults aged 75 and over. Household car availability was grouped into households with no car, one car, and two or more cars. Educational attainment was grouped into no formal or primary education, secondary education, and tertiary education. The percentage still in education was retained as a separate demographic covariate, rather than being included in the educational attainment composition, because it was calculated using a different denominator.

To account for the compositional nature of these variables, the isometric log-ratio transformation was applied to each composition. The ILR transformation converts a composition with $D$ parts into $D - 1$ real-valued coordinates that can be used in standard regression models [@egozcue2003ilr]. These coordinates represent balances between groups of compositional parts rather than isolated changes in individual percentages. For example, an ILR coordinate for age composition may represent a contrast between younger and older age groups, rather than the independent effect of one age group alone.

For a composition $\mathbf{x}_i = (x_{i1}, x_{i2}, ..., x_{iD})$ in Census Small Area (i), the transformation can be written generally as:

$$
\mathbf{z}_i = ilr(\mathbf{x}_i)
$$

where $\mathbf{z}_i$ is the set of (D - 1) ILR coordinates. These transformed coordinates are unconstrained and can therefore be included as regression covariates without the perfect collinearity that would arise from including all raw percentage components directly.

Exact zero values cannot be used directly in log-ratio transformations because the logarithm of zero is undefined. Before applying the ILR transformation, exact zero percentages were therefore replaced with a small positive value of 0.001. In practice, exact zeros were uncommon and occurred mainly in the car-availability and educational-attainment compositions. This allowed the transformation to be computed while making only a minimal change to the original percentage values.

The ILR transformation was implemented in R using the `compositions` package [@vandenboogaart2008compositions]. The resulting ILR-transformed demographic variables were included as covariates in the Bayesian beta regression model. This allowed the model to adjust for demographic differences between urban Census Small Areas while avoiding the collinearity problems associated with raw percentage variables. The original percentage variables were retained for descriptive summaries, but the ILR-transformed variables were used in the regression model.

## Statistical Model: Bayesian Beta Regression
The main statistical model was used to estimate the association between school accessibility and active school travel in urban Census Small Areas. The outcome variable was the proportion of pupils travelling to school by active modes, defined as walking or cycling. Because this outcome is continuous and bounded between 0 and 1, beta regression was used. Beta regression is appropriate for modelling rates and proportions that lie within the open interval ((0, 1)), and allows the expected value of the outcome to be related to explanatory variables through a regression structure [@ferrari2004beta].

The analysis was restricted to urban Census Small Areas because the main research question focuses on active school travel in settings where walking and cycling are more feasible. Before modelling, Census Small Areas with missing required model inputs were removed. In addition, areas with total population greater than 900 and areas with fewer than five residents aged 3–18 were excluded. These exclusions were applied to reduce the influence of unusually large Census Small Areas and unstable school-age population denominators. The final urban modelling dataset therefore included Census Small Areas with valid active-travel outcomes, valid school accessibility measures, and sufficient population information.

The active-travel percentage was converted to a proportion before modelling. Since the beta distribution is defined only on the open interval ((0, 1)), any exact zero or one values were adjusted slightly so that all observations lay strictly within this interval. This type of transformation is commonly used when applying beta regression to proportional outcomes that may include boundary values [@smithson2006better].

The model was fitted using a Bayesian beta regression framework. For Census Small Area (i), the outcome was modelled as:

$$
y_i \sim \mathrm{Beta}(\mu_i, \phi)
$$

where $y_i$ is the active-travel proportion, $\mu_i$ is the expected active-travel proportion, and $\phi$ is the beta precision parameter. Under this parameterisation, $E(y_i) = \mu_i$. Larger values of $\phi$ indicate that observations are more tightly concentrated around the expected proportion, while smaller values indicate greater variability.

A logit link function was used to relate the expected active-travel proportion to the explanatory variables:

$$
\mathrm{logit}(\mu_i) =
\alpha +
\sum_{j=1}^{p} \beta_j Z_{ij}
+
f_1(K_i)
+
f_2(D_i)
$$

where $\alpha$ is the intercept, $Z_{ij}$ represents the set of standardised demographic covariates for Census Small Area $i$, $K_i$ is the normalised kriging-based school-place density measure, and $D_i$ is the secondary spatial density measure of predicted school places per square kilometre. The demographic covariates included the ILR-transformed age composition, household car-availability composition, and educational-attainment composition, together with the additional percentage still in education covariate where included. These variables were used to adjust for differences in demographic structure across urban Census Small Areas.

The same model structure was fitted twice, once for each accessibility measure. In the first model, the accessibility exposure was the kriging-based school-place density measure described above. In the second model, this was replaced by the walking accessibility measure, the log-transformed count of school places reachable on foot within 30 minutes. All demographic covariates and other model settings were kept identical between the two models, so that the kriging-based and walking-based measures could be compared on the same outcome and the same adjustment set. These are referred to as the kriging model and the walking model in the results.

The terms $f_1(K_i)$ and $f_2(D_i)$ represent natural cubic spline functions. Natural cubic splines were used because the relationship between school accessibility and active travel may not be linear. For example, increases in school accessibility may be strongly associated with active travel at low levels of accessibility, but the association may become weaker or flatten at higher levels. Using splines allows the model to capture this type of non-linear relationship without assuming a single constant effect across the full range of the accessibility variables. Three degrees of freedom were used for each spline, providing a flexible but controlled curve and reducing the risk of overfitting.

The normalised kriging-based accessibility variable was expressed as predicted school-place availability per 100 children aged 3–18. This variable captures school-place availability relative to local school-age population demand. The additional school-place density per square kilometre variable was included to capture the spatial concentration of predicted school places independently of the local child population denominator. Including both variables allowed the model to distinguish between relative school-place availability and broader spatial concentration of school places.

To account for differences in the reliability of active-travel percentages across Census Small Areas, the model used weights based on the active-travel denominator. Census Small Areas with larger numbers of pupils with a stated travel mode provide more stable estimates of active-travel proportion than areas with smaller denominators. The weights were calculated as the square root of the active-travel denominator and then normalised to have an average value of one in the urban modelling sample. This gave more influence to areas with more reliable active-travel estimates while avoiding excessive dominance by the largest areas.

Weakly informative priors were used for the model parameters. Regression coefficients were assigned normal priors centred at zero, which regularised the model by discouraging unrealistically large effects while still allowing the data to inform the posterior estimates. The intercept was assigned a wider normal prior on the logit scale. The beta precision parameter $\phi$ was assigned a positive prior to ensure that it remained greater than zero. These priors were chosen to provide regularisation without imposing strong assumptions about the direction or size of the accessibility effects.

The model was fitted using the `brms` package in R, which fits Bayesian regression models using Stan [@burkner2017brms]. Posterior sampling was carried out using Hamiltonian Monte Carlo with the No-U-Turn Sampler, as implemented in Stan [@hoffman2014nuts]. Four Markov chains were run, and convergence was assessed using standard Bayesian diagnostics, including $\hat{R}$, effective sample size, trace plots, divergent transitions, and posterior predictive checks. Values of $\hat{R}$ close to 1 and adequate effective sample sizes were taken as evidence that the chains had mixed well [@vehtari2021rhat].

Because spline coefficients are not directly interpretable as simple one-unit changes, the results were interpreted using posterior predictions and marginal effect plots on the response scale. These plots show how the expected active-travel percentage changes across the observed range of the accessibility variables. Interpretation was based on posterior means and credible intervals, rather than only on raw regression coefficients. Posterior predictive checks were used to assess whether the fitted model could reproduce the observed distribution of active travel.

# Results
The outcome variable in this study is the percentage of pupils travelling to school by active modes (walking or cycling) in each Census Small Area. Figure 7 shows how this active-travel percentage is distributed, with urban and rural areas shown separately.

The two distributions are clearly different. In urban areas, active travel is common and varies widely between areas: the distribution is spread out, with a mean and median both close to 38%, and many areas reaching much higher levels. In rural areas, active travel is much lower, with most areas clustered near the bottom of the scale (a median of around 5%) and only a small number of areas reaching higher values. The mean for rural areas (around 9%) is pulled up by this small tail, which is why it sits above the median.

This difference is what would be expected: in rural areas, longer distances between homes and schools make walking and cycling less feasible, so most children travel by car or bus. In urban areas, schools are closer and the street network is more connected, so active travel is a realistic option for many more children. This urban–rural contrast is the main reason the statistical model was restricted to urban Census Small Areas, where active school travel is genuinely possible and the accessibility measures are most meaningful.

```{r active-travel-urban-rural, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, out.width="100%", fig.cap="Active-travel percentage across Census Small Areas, urban vs rural. White point = median, black point = mean. Active travel is higher and more variable in urban areas, low in most rural areas."}

library(tidyverse)

# active-travel outcome + urban/rural flag
at_src <- readRDS("D:/Final-Project/DATA/final_df_complete.rds") %>%
  sf::st_drop_geometry() %>%
  filter(is.finite(perc_active), !is.na(urban_rural)) %>%
  filter(perc_active <= quantile(perc_active, 0.99, na.rm = TRUE)) %>%
  select(urban_rural, perc_active)

# summary points
at_summary <- at_src %>%
  group_by(urban_rural) %>%
  summarise(mean_at = mean(perc_active, na.rm = TRUE),
            median_at = median(perc_active, na.rm = TRUE), .groups = "drop")

ggplot(at_src, aes(x = urban_rural, y = perc_active)) +
  geom_violin(fill = "lightblue", colour = "lightblue", trim = TRUE) +
  geom_point(data = at_summary, aes(y = median_at),
             colour = "white", size = 3, shape = 21, fill = "white", stroke = 0.8) +
  geom_point(data = at_summary, aes(y = mean_at),
             colour = "black", size = 2.5) +
  coord_flip() +
  labs(
    title = "Active travel to school, urban vs rural",
    subtitle = "White point = median, black point = mean; below 99th percentile",
    x = NULL,
    y = "Active travel (% of pupils walking or cycling)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

## Spatial Distribution of Schools
The 2024/25 school dataset was first explored spatially to understand the distribution of school locations across Ireland. Each school was represented as a point using its latitude and longitude coordinates, and points were mapped against the national boundary. This provides an initial descriptive view of where schools are concentrated before constructing the kriging-based and travel-time accessibility measures.

```{r spatial-distribution-schools, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of schools across Ireland in the 2024/25 mainstream school dataset. Points represent school locations and are coloured by school level."}
library(readxl)
library(dplyr)
library(sf)
library(ggplot2)
library(rnaturalearth)

schools_df <- read_excel(
  "D:/Final-Project/DATA/Data_on_Individual_Schools_Mainstream_2024_25.xlsx",
  sheet = 2
) %>%
  rename(
    Roll_Number = `Roll Number`,
    longitude = `School Longitude`,
    latitude = `School Latitude`,
    AcademicYear = `Academic Year (Enrolment)`,
    SchoolName = `Official Name`
  )

schools_df_clean <- schools_df %>%
  filter(!is.na(latitude), !is.na(longitude))

schools_sf <- st_as_sf(
  schools_df_clean,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
)

ireland <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Ireland")

ggplot() +
  geom_sf(
    data = ireland,
    fill = "grey95",
    colour = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data = schools_sf,
    aes(colour = `School Level`),
    size = 0.7,
    alpha = 0.75
  ) +
  labs(
    title = "Spatial distribution of schools in Ireland",
    subtitle = "Mainstream schools, 2024/25",
    colour = "School level"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```
The map shows that schools are distributed across the country, with visible clustering in and around larger urban areas. This spatial pattern is important for the later analysis because school accessibility is not determined only by the number of schools nationally, but by how school places are distributed relative to where children live. The mapped school locations therefore provide the spatial input for both the kriging-based school-place density measure and the routing-based travel-time accessibility analysis.

## Walking Accessibility   
This section describes the walking accessibility measure across Census Small Areas (CSAs). The measure counts how many school places can be reached on foot within 30 minutes from each CSA, expressed per 100 children aged 3–18. It is based on routing-based travel time, which uses the real road and path network rather than straight-line distance, so it reflects the routes a child would actually use and captures the effect of any barriers in the network.

Figure 8 maps the walking accessibility measure, with its distribution shown alongside. The island of Ireland boundary is drawn beneath the map for geographical context. The values are strongly right-skewed: a small number of dense urban CSAs can reach many thousands of school places, while most CSAs reach far fewer. The median walking accessibility is about 920 places per 100 children, but the highest values reach into the tens of thousands. Because of this skew, the map colour scale uses a square-root transformation so that differences at the lower end of the distribution remain visible, which a linear scale would hide. The map shows that walking accessibility is highest in and around the larger urban areas, where schools and street networks are dense, and decreases towards the edges of towns and along the coast. Some CSAs on peninsulas, islands, and estuary edges have very low or zero walking accessibility even though schools are not far away in straight-line distance. This pattern is important for the comparison that follows, because it suggests that barriers in the network, rather than distance alone, affect how reachable schools really are. The distribution shown beside the map confirms the strong right skew, including a group of CSAs at or near zero walking accessibility.

```{r walking-map-and-distribution, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, out.width="100%", fig.cap="Walking accessibility across Census Small Areas (left) and its distribution (right): school places reachable on foot within 30 minutes per 100 children aged 3-18. Colour scale is square-root transformed so lower values stay visible."}

library(tidyverse)
library(sf)
library(scales)
library(patchwork)

walk_src <- readRDS("D:/Final-Project/DATA/final_df_complete.rds") %>%
  sf::st_drop_geometry() %>%
  mutate(
    SA_PUB2022 = as.character(SA_PUB2022),
    access_walk_per_100 = 100 * (access_walk_30 / pop_3_18)
  ) %>%
  filter(pop_3_18 >= 5) %>%
  select(SA_PUB2022, access_walk_per_100)

walk_map_sf <- final_sf %>%
  mutate(SA_PUB2022 = as.character(SA_PUB2022)) %>%
  left_join(walk_src, by = "SA_PUB2022") %>%
  st_make_valid() %>%
  mutate(
    access_walk_cap = pmin(
      access_walk_per_100,
      quantile(access_walk_per_100, 0.95, na.rm = TRUE)
    )
  )

p_map <- ggplot() +
  geom_sf(data = ireland_boundary, fill = "grey97", colour = "grey70", linewidth = 0.2) +
  geom_sf(data = walk_map_sf, aes(fill = access_walk_cap), colour = NA) +
  geom_sf(data = ireland_boundary, fill = NA, colour = "grey40", linewidth = 0.15) +
  scale_fill_viridis_c(name = "Walking access\nper 100 children",
                       trans = "sqrt", labels = label_number(big.mark = ",")) +
  labs(title = "Walking accessibility to school places") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_hist <- walk_map_sf %>%
  st_drop_geometry() %>%
  filter(is.finite(access_walk_per_100)) %>%
  ggplot(aes(x = access_walk_per_100)) +
  geom_histogram(bins = 40, fill = "lightblue", colour = "white") +
  scale_x_sqrt(labels = label_number(big.mark = ",")) +
  labs(title = "Distribution", x = "Walking access per 100 children",
       y = "Number of Census Small Areas") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_map + p_hist
```
## How the Two Measures Compare                  
Having described each accessibility measure separately, this section compares them directly. The comparison is restricted to urban CSAs, matching the scope of the later statistical model. Figure 10 plots the kriged school-place density against the walking accessibility measure, with both expressed per 100 children aged 3–18 and shown on logarithmic scales. Each point is coloured by the difference in rank between the two measures, so that areas ranked similarly by both measures appear pale, while areas ranked very differently appear strongly coloured.

Overall, the two measures are positively related: CSAs with higher kriged density tend also to have higher walking accessibility, and the cloud of points slopes upward. However, the relationship is loose rather than tight, and there is a clear spread of CSAs that the two measures rank differently. A band of CSAs sits along the bottom of the plot at zero walking accessibility despite having moderate or high kriged density. These are areas where schools appear close in straight-line terms but cannot be reached on foot within 30 minutes.

Figure 11 summarises this disagreement using the difference in rank between the two measures. The distribution is broadly symmetric and centred near zero, which confirms that neither measure systematically ranks areas higher than the other. At the same time, the spread on either side of zero shows that the two measures often disagree for individual CSAs. In other words, the kriging and walking measures agree on average but diverge in specific places. The next section examines these divergent areas in more detail, since they point to a consistent explanation.

```{r kriging-vs-walking-scatter, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="100%", fig.cap="Kriged school-place density against walking accessibility for urban Census Small Areas, both per 100 children on log scales. Points are coloured by rank difference between the two measures; the dashed line marks equality."}

library(tidyverse)
library(sf)
library(scales)

krig_vals <- final_df %>%
  select(SA_PUB2022, krig_value_per_100_3_18) %>%
  mutate(SA_PUB2022 = as.character(SA_PUB2022))

walk_vals <- readRDS("D:/Final-Project/DATA/final_df_complete.rds") %>%
  sf::st_drop_geometry() %>%
  mutate(
    SA_PUB2022 = as.character(SA_PUB2022),
    access_walk_per_100 = 100 * (access_walk_30 / pop_3_18)
  ) %>%
  filter(pop_3_18 >= 5, urban_rural == "Urban") %>%
  select(SA_PUB2022, access_walk_per_100)

compare_df <- krig_vals %>%
  inner_join(walk_vals, by = "SA_PUB2022") %>%
  filter(is.finite(krig_value_per_100_3_18), is.finite(access_walk_per_100)) %>%
  mutate(
    rank_krig = rank(krig_value_per_100_3_18),
    rank_walk = rank(access_walk_per_100),
    rank_diff = rank_krig - rank_walk
  )

ggplot(compare_df,
       aes(x = krig_value_per_100_3_18, y = access_walk_per_100, colour = rank_diff)) +
  geom_point(alpha = 0.55, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black") +
  scale_x_log10(labels = label_number(big.mark = ",")) +
  scale_y_log10(labels = label_number(big.mark = ",")) +
  scale_colour_gradient2(name = "Kriging - walk\nrank difference",
                         low = "#2166AC", mid = "grey85", high = "#B2182B", midpoint = 0) +
  labs(x = "Kriged school-place density (per 100 children, log scale)",
       y = "Walking accessibility (per 100 children, log scale)",
       title = "Kriging vs walking accessibility, urban Census Small Areas") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))
```

```{r rank-difference-histogram, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5, out.width="100%", fig.cap="Distribution of the rank difference between the kriging and walking measures across urban Census Small Areas. Zero (dashed line) means the two measures rank an area equally; positive values mean kriging ranks it higher than walking."}

ggplot(compare_df, aes(x = rank_diff)) +
  geom_histogram(bins = 40, fill = "grey40", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") +
  labs(x = "Rank difference (kriging rank - walking rank)",
       y = "Number of Census Small Areas",
       title = "Distribution of the rank difference between the two measures") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))
```

## Where the Measures Diverge: Severance                    
The comparison in the previous section showed that the kriging and walking measures agree on average but disagree for individual CSAs. This section looks more closely at the areas where they diverge most. The most informative of these are CSAs where the kriged density is moderate or high, suggesting schools are nearby in straight-line terms, but the walking accessibility is zero, meaning no school places can actually be reached on foot within 30 minutes. This pattern is consistent with severance: physical barriers such as motorways, rivers, and stretches of water that cut an area off from schools that are close by but not reachable on foot.

Figure 12 highlights these CSAs. Across the urban CSAs, a group of areas with above-median kriged density also have zero walking accessibility, and they are circled in the plot. Table 1 lists the strongest of these cases, ordered by kriged density. In each of these CSAs, the kriging measure ranks the area as having good school-place availability, while the walking measure ranks it among the lowest, because the network distance to any school exceeds a 30-minute walk.
```{r severance-scatter, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="100%", fig.cap="Urban Census Small Areas with zero walking accessibility but above-median kriged density are circled: schools are close in straight-line terms but not reachable on foot in 30 minutes. Zero-walk areas are plotted at one for the log axis."}

library(tidyverse)
library(scales)

krig_mid <- median(compare_df$krig_value_per_100_3_18, na.rm = TRUE)

severance <- compare_df %>%
  filter(access_walk_per_100 == 0, krig_value_per_100_3_18 >= krig_mid)

plot_df <- compare_df %>%
  mutate(walk_plot = ifelse(access_walk_per_100 == 0, 1, access_walk_per_100))

sev_plot <- severance %>% mutate(walk_plot = 1)

ggplot(plot_df, aes(x = krig_value_per_100_3_18, y = walk_plot)) +
  geom_point(aes(colour = rank_diff), alpha = 0.45, size = 1) +
  geom_point(data = sev_plot, shape = 21, size = 2.5, stroke = 0.7,
             colour = "black", fill = NA) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey40") +
  scale_x_log10(labels = label_number(big.mark = ",")) +
  scale_y_log10(labels = label_number(big.mark = ",")) +
  scale_colour_gradient2(name = "Kriging - walk\nrank difference",
                         low = "#2166AC", mid = "grey85", high = "#B2182B", midpoint = 0) +
  labs(x = "Kriged school-place density (per 100 children, log scale)",
       y = "Walking accessibility (per 100 children, log scale)",
       title = "Schools nearby but none walkable: zero-walk areas circled") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))
```

```{r severance-table, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)

severance %>%
  arrange(desc(krig_value_per_100_3_18)) %>%
  head(15) %>%
  transmute(
    `Census Small Area` = SA_PUB2022,
    `Kriged density (per 100)` = round(krig_value_per_100_3_18, 0),
    `Walking access (per 100)` = round(access_walk_per_100, 0),
    `Kriging rank` = round(rank_krig, 0),
    `Walking rank` = round(rank_walk, 0)
  ) %>%
  kable(caption = "The fifteen urban Census Small Areas with the highest kriged school-place density but zero walking accessibility.")

```

```{r jacobs-island-severance, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6, out.width="100%", fig.cap="Severance near Cork. Left: Jacob's Island in the wider Cork area, with schools in yellow. Right: the highlighted Census Small Area (047132051), cut off from nearby schools by the Mahon estuary and the South Ring Road, so none can be reached on foot in 30 minutes."}

library(png)
library(grid)
library(patchwork)

img1 <- rasterGrob(readPNG("D:/Final-Project/images/JB1.png"), interpolate = TRUE)
img2 <- rasterGrob(readPNG("D:/Final-Project/images/JB3.png"), interpolate = TRUE)

(wrap_elements(img1) + wrap_elements(img2)) +
  plot_annotation(
    title = "Severance in action: Jacob's Island, Cork",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )


```
A clear example is the CSA on Jacob's Island in Cork, shown in Figure 13. The wider view on the left shows that schools (in yellow) are present throughout the Cork area, including several within a short straight-line distance of Jacob's Island. The zoomed view on the right shows the CSA itself, which sits on a spit of land surrounded by the Mahon estuary and cut off from the city by the South Ring Road. Although schools are close as the crow flies, the only routes available involve long detours by road, so no school can be reached on foot within 30 minutes and the walking accessibility for this area is zero. The kriging measure, which is based on straight-line distance, does not capture this barrier and therefore records the area as having reasonable school-place availability.

The same broad pattern can be seen at the national scale in the walking accessibility map (Figure 9), where the lowest-accessibility CSAs cluster along the coast, on peninsulas, and around estuaries. This suggests that severance is a recurring feature rather than an isolated case, and that straight-line measures of school-place availability can overstate how reachable schools actually are in areas affected by physical barriers. It should be noted that this is a consistent pattern identified by comparing the two measures, rather than a formally tested effect, but the geography of the divergent areas points clearly towards severance as the explanation.

```{r bivariate-map, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=9, out.width="100%", fig.cap="Bivariate map shading each Census Small Area by both kriged density and walking accessibility at once. Areas high in both (dark) cluster in the cities; areas with high density but low walking access (red) cluster along the coast and estuaries, showing severance at national scale."}

library(tidyverse)
library(sf)
library(biscale)
library(cowplot)

# both measures joined to geometry (all CSAs, population filter only)
walk_src <- readRDS("D:/Final-Project/DATA/final_df_complete.rds") %>%
  sf::st_drop_geometry() %>%
  mutate(SA_PUB2022 = as.character(SA_PUB2022),
         access_walk_per_100 = 100 * (access_walk_30 / pop_3_18)) %>%
  filter(pop_3_18 >= 5) %>%
  select(SA_PUB2022, access_walk_per_100)

bi_sf <- final_sf %>%
  mutate(SA_PUB2022 = as.character(SA_PUB2022)) %>%
  inner_join(walk_src, by = "SA_PUB2022") %>%
  filter(is.finite(krig_value_per_100_3_18), is.finite(access_walk_per_100)) %>%
  st_make_valid()

# classify into a 3x3 bivariate grid (quantile breaks)
bi_sf <- bi_class(bi_sf,
                  x = krig_value_per_100_3_18,
                  y = access_walk_per_100,
                  style = "quantile", dim = 3)

map <- ggplot() +
  geom_sf(data = ireland_boundary, fill = "grey97", colour = "grey70", linewidth = 0.2) +
  geom_sf(data = bi_sf, aes(fill = bi_class), colour = NA, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  labs(title = "Two accessibility measures at once",
       subtitle = "Kriged density and walking accessibility, Census Small Areas") +
  bi_theme() +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10))

legend <- bi_legend(pal = "GrPink", dim = 3,
                    xlab = "Higher kriged density",
                    ylab = "Higher walking access",
                    size = 8)

ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +
  draw_plot(legend, 0.02, 0.02, 0.22, 0.22)
```
The bivariate map in Figure 14 shows both accessibility measures together across the whole country. Each Census Small Area is shaded by its kriged school-place density and its walking accessibility at once, using the colour grid shown in the corner. Areas scoring high on both measures appear dark and are concentrated in the cities — Dublin, Cork, Galway, and Limerick stand out as dark clusters where schools are both densely distributed and easy to reach on foot. Areas scoring high on kriged density but low on walking accessibility appear in the red tones; these are the severance areas, where schools are close in straight-line terms but not reachable on foot.

The red severance areas are most common around the coast, on peninsulas, and in areas broken up by water, including stretches of the west and south-west coast and the Donegal coastline. This matches the pattern seen in the walking accessibility map, and supports the interpretation that severance is driven by physical geography: where the land is divided by estuaries, rivers, or the sea, straight-line proximity to schools does not translate into walkable access. Seen at national scale, the divergence between the two measures is a consistent, geographically patterned feature rather than a handful of isolated cases.

## Model Results — Urban Areas                            
Two Bayesian beta regression models were fitted for urban CSAs: one using the kriged school-place density measure and one using the walking accessibility measure, each adjusted for the same set of demographic covariates. Before interpreting the estimated effects, the models were checked for convergence and fit.

Figure 15 shows the estimated coefficients for both models, with their 95% credible intervals. The demographic covariates are shown on the ILR-transformed and standardised scale, so their coefficients are not directly interpretable as simple percentage changes, but the plot shows which covariates are more clearly associated with active travel. The clearest difference between the two models is in the accessibility spline terms. In the walking model, the spline terms for walking accessibility have credible intervals clearly separated from zero, indicating a strong association with active travel. In the kriging model, the corresponding terms sit closer to zero with wider intervals, indicating a weaker and less certain association.

Figure 16 shows posterior predictive checks for both models. In each panel, the observed distribution of the active-travel proportion (the dark line) sits within the spread of the simulated distributions drawn from the fitted model (the lighter lines), which indicates that both models reproduce the observed pattern of active travel reasonably well.

Table 2 reports the convergence diagnostics for both models. In both cases the maximum Rhat value is very close to one (1.002 for the kriging model and 1.001 for the walking model), the effective sample sizes are in the thousands, and there were no divergent transitions. Together these indicate that the Markov chains mixed well and that the posterior estimates are reliable. Table 3 provides summary statistics for the main variables used in the model. Because the accessibility effects are modelled with splines and are easier to read on the response scale, they are presented as marginal-effect curves in the next section.
            
```{r coefficient-plots-both, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, out.width="100%", fig.cap="Posterior coefficient estimates (means and 95% credible intervals) for the kriging model (left) and walking model (right). Intervals that do not cross zero indicate covariates more clearly associated with active travel."}

library(brms)
library(tidyverse)
library(patchwork)

model_urban_krig <- readRDS("D:/Final-Project/model/models/model_urban_krig_splines.rds")
model_urban_walk <- readRDS("D:/Final-Project/model/models/model_urban_walk.rds")

p_krig <- mcmc_plot(model_urban_krig, type = "intervals", prob = 0.95) +
  labs(title = "Kriging model") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_walk <- mcmc_plot(model_urban_walk, type = "intervals", prob = 0.95) +
  labs(title = "Walking model") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_krig + p_walk
```

```{r posterior-predictive-both, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=5, out.width="100%", fig.cap="Posterior predictive checks for the kriging model (left) and walking model (right). The dark line is the observed active-travel distribution; lighter lines are draws from the fitted model. Close agreement indicates a good fit."}

library(patchwork)

p_pp_krig <- pp_check(model_urban_krig, ndraws = 100) +
  labs(title = "Kriging model") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_pp_walk <- pp_check(model_urban_walk, ndraws = 100) +
  labs(title = "Walking model") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_pp_krig + p_pp_walk
```
     
```{r convergence-table-both, echo=FALSE, message=FALSE, warning=FALSE}
library(brms)
library(posterior)
library(tidyverse)
library(knitr)

model_urban_krig <- readRDS("D:/Final-Project/model/models/model_urban_krig_splines.rds")
model_urban_walk <- readRDS("D:/Final-Project/model/models/model_urban_walk.rds")

conv_row <- function(model, name) {
  ds <- summarise_draws(model)
  np <- nuts_params(model)
  tibble(
    Model = name,
    `Max Rhat` = round(max(ds$rhat, na.rm = TRUE), 3),
    `Min bulk ESS` = round(min(ds$ess_bulk, na.rm = TRUE), 0),
    `Min tail ESS` = round(min(ds$ess_tail, na.rm = TRUE), 0),
    `Divergences` = sum(np$Value[np$Parameter == "divergent__"])
  )
}

bind_rows(
  conv_row(model_urban_krig, "Kriging"),
  conv_row(model_urban_walk, "Walking")
) %>%
  kable(caption = "Convergence diagnostics for both models. Rhat values near 1, large effective sample sizes (ESS), and no divergent transitions indicate the chains converged well.")
```

```{r summary-statistics-table, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)

df_model <- readRDS("D:/Final-Project/model/data2/df_urban_model.rds")

# pick the key model variables (response + accessibility + a few covariates)
summary_vars <- df_model %>%
  transmute(
    `Active travel (%)`               = perc_active,
    `Kriged density (per 100)`        = krig_value_per_100_3_18,
    `Walking access (per 100)`        = access_walk_per_100,
    `School-place density (per km2)`  = places_density_km2,
    `Population aged 3-18`            = pop_3_18,
    `Total population`                = total_pop
  )

summary_vars %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "value") %>%
  group_by(Variable) %>%
  summarise(
    Mean   = round(mean(value, na.rm = TRUE), 1),
    Median = round(median(value, na.rm = TRUE), 1),
    SD     = round(sd(value, na.rm = TRUE), 1),
    Min    = round(min(value, na.rm = TRUE), 1),
    Max    = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  kable(caption = "Summary statistics for the main variables used in the urban modelling dataset. Values are shown for the Census Small Areas included in the model after filtering.")
```

## Effect of Accessibility on Active Travel
The marginal-effect curves show how the predicted percentage of active travel changes across the range of each accessibility measure, with all other covariates held at their average. Because the spline coefficients are not directly interpretable, these curves give a clearer picture of each measure's effect on the response scale. Figure 17 shows the two curves. The two measures entered their models on different scales (the kriging measure on its original scale and the walking measure on a log10 scale), so the curves should be read individually rather than compared by their position on the horizontal axis. They can, however, be compared on the vertical axis, since both predict the same outcome.
The kriging curve (left) is gentle and slightly non-monotonic. Predicted active travel changes only modestly across the range of kriged density, moving within a band of roughly 41% to 44%, and the credible interval is wide. This indicates that kriged school-place density has only a small and uncertain association with active travel once demographic differences are accounted for.

The walking curve (right) is much steeper and has a clear S-shape. Predicted active travel rises from around 15% at the lowest levels of walking accessibility to around 60% at the highest, and the credible interval is narrow throughout. This indicates a strong and fairly certain association: areas where more school places can actually be reached on foot have substantially higher active travel.

Comparing the two curves on the vertical axis, the walking measure is associated with a much larger change in active travel than the kriging measure, and with greater certainty. This is consistent with the severance pattern described earlier: what matters for active travel is not simply how close schools are in straight-line terms, but whether they can actually be reached on foot. The walking measure captures this directly, which may explain why it is more strongly associated with active travel than the straight-line kriging measure.

```{r marginal-effects-both, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6, out.width="100%", fig.cap="Marginal effect of each accessibility measure on predicted active-travel percentage, with other covariates held at their average. Left: kriged density (original scale). Right: walking accessibility (log10 scale). The curves should be read individually, not compared by horizontal position."}

library(brms); library(tidyverse); library(patchwork); library(splines)

model_urban_krig <- readRDS("D:/Final-Project/model/models/model_urban_krig_splines.rds")
model_urban_walk <- readRDS("D:/Final-Project/model/models/model_urban_walk.rds")
df_urban_model   <- readRDS("D:/Final-Project/model/data2/df_urban_model.rds")

# helper: hold all numeric predictors at their mean, vary one column
make_newdata <- function(df, vary_col, grid) {
  template <- df %>% summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
  nd <- template[rep(1, length(grid)), ]
  nd[[vary_col]] <- grid
  nd$prop_weight <- 1
  nd
}

# kriging
kg <- seq(quantile(df_urban_model$krig_value_per_100_3_18, .01, na.rm=TRUE),
          quantile(df_urban_model$krig_value_per_100_3_18, .99, na.rm=TRUE), length.out=100)
ndk <- make_newdata(df_urban_model, "krig_value_per_100_3_18", kg)
pk  <- fitted(model_urban_krig, newdata = ndk, scale="response", probs=c(.025,.975))
plot_krig <- bind_cols(ndk, as_tibble(pk)) %>%
  mutate(across(c(Estimate,Q2.5,Q97.5), ~.x*100))

# walking
wg <- seq(quantile(df_urban_model$log10_walk_30, .01, na.rm=TRUE),
          quantile(df_urban_model$log10_walk_30, .99, na.rm=TRUE), length.out=100)
ndw <- make_newdata(df_urban_model, "log10_walk_30", wg)
pw  <- fitted(model_urban_walk, newdata = ndw, scale="response", probs=c(.025,.975))
plot_walk <- bind_cols(ndw, as_tibble(pw)) %>%
  mutate(across(c(Estimate,Q2.5,Q97.5), ~.x*100))

# kriging on its own (linear) scale
p_krig <- ggplot(plot_krig, aes(krig_value_per_100_3_18, Estimate)) +
  geom_ribbon(aes(ymin=Q2.5, ymax=Q97.5), alpha=.2) + geom_line(linewidth=1) +
  labs(title="Kriging accessibility effect",
       x="Kriged school-place density (per 100 children)",
       y="Predicted active travel (%)") +
  theme_minimal() + theme(plot.title=element_text(face="bold"))

# walking on its log10 scale
p_walk <- ggplot(plot_walk, aes(log10_walk_30, Estimate)) +
  geom_ribbon(aes(ymin=Q2.5, ymax=Q97.5), alpha=.2) + geom_line(linewidth=1) +
  scale_x_continuous(breaks=c(0,1,2,3,4),
                     labels=c("1","10","100","1,000","10,000")) +
  labs(title="Walking accessibility effect",
       x="Walking accessibility (per 100 children, log scale)",
       y="Predicted active travel (%)") +
  theme_minimal() + theme(plot.title=element_text(face="bold"))

p_krig + p_walk
```

# Discussion

## Interpretation of Results
The central finding of this study is that how schools are reached matters more than how close they are in straight-line terms. The two accessibility measures agreed on average but diverged in many individual CSAs, and the areas where they diverged most pointed consistently to severance, where physical barriers such as motorways, rivers, and water separate an area from schools that are nearby as the crow flies. This interpretation is supported by the statistical model. The walking accessibility measure, which is based on actual network travel time, had a strong and fairly certain association with active travel, while the kriged density measure, which is based on straight-line distance, had only a small and uncertain association. Taken together, the comparison and the model suggest that straight-line measures of school-place availability can overstate how reachable schools really are, and that network-based measures capture the practical opportunity for active travel more faithfully.

These findings connect back to the "postcode lottery" idea that motivated the study. They suggest that unequal opportunities for active school travel are not only about where schools are located, but also about whether the local network allows children to reach them on foot. Two areas with similar straight-line access to schools can have very different walking opportunities if one is cut off by a barrier. This points to the potential importance of local 
infrastructure, such as pedestrian crossings and routes across or around barriers, as well as the broader distribution of school places, in shaping opportunities for active school travel.

## Limitations
A number of limitations should be kept in mind when interpreting these results. The outcome data are restricted by the way school travel information is released in official statistics. The Census data provide aggregated travel-to-education counts for Census Small Areas, rather than individual-level journeys between home and school. As a result, the analysis cannot observe the actual school attended by each pupil, the exact route taken, or whether pupils travel to primary or post-primary schools. The model therefore estimates area-level associations between school accessibility and active travel, rather than individual travel behaviour, and the results should be interpreted as evidence of spatial association, not as proof that improved accessibility directly causes higher walking or cycling rates.

The active-travel outcome is also not separated by school type. Primary and post-primary pupils may have different travel distances and school-choice patterns: younger children may depend on nearby primary schools, while older students may travel further to fewer, more specialised post-primary schools. Because the Census outcome combines both groups, the analysis cannot distinguish these patterns. Future work could repeat the analysis separately for primary and post-primary schools where the data allow, although this may be limited by the availability of travel-mode outcomes disaggregated by school type at fine spatial scales.

The school accessibility measures use enrolment as a proxy for school places. Enrolment provides a consistent measure of the relative size and spatial distribution of schools, but it does not measure spare capacity, catchment areas, admissions policies, or whether a place is realistically available to a child living in a particular area. This limitation is especially relevant where schools are oversubscribed or where school choice is shaped by factors other than distance.

The kriging-based accessibility measure captures the spatial distribution of school places, but it does not account for barriers in the transport network, route safety, traffic conditions, or the quality of walking infrastructure. The routing-based travel-time measure addresses this limitation more directly by using the road and path network, but it still depends on the assumed walking speed and on the completeness and accuracy of OpenStreetMap data.
Finally, the workflow could be repeated using settlement-level data as a complementary analysis. Settlement-level travel tables can describe travel behaviour for whole towns or urban areas, rather than only Census Small Areas. This may provide a broader view of how school provision and active travel relate at the town scale, and could be useful for comparing with previous work on Irish settlements. However, settlement-level data are less spatially detailed than Census Small Area data, so they would be better suited as a complementary robustness or contextual analysis rather than a replacement for the main local-scale model.


## Comparison with Previous Work
The motivation for this work came from Mancini's research on adolescent active transport in Ireland, which argued that differences in active travel between areas are linked to the geography of school provision rather than simply to population size [@mancini2025census; @mancini2024rte]. The findings here are broadly consistent with that argument, but they add a spatial dimension to it. Where Mancini focused on the presence and type of school provision, the results presented here show that even where provision appears adequate in straight-line terms, active travel can still be limited if schools are not reachable on foot through the local network.

In this sense, the present analysis complements rather than contradicts the earlier work. It supports the idea that active school travel is shaped by the geography of provision, while highlighting that the connectivity of the network is an additional and important part of that geography. The "postcode lottery" of active travel may therefore depend not only on whether school places exist nearby, but on whether the surrounding network allows children to walk to them. Working at the finer Census Small Area scale, rather than the settlement level, made it possible to detect the local severance effects that a coarser analysis would miss.

# Conclusions
The aim of this work was to examine whether school-place accessibility is associated with active school travel across Ireland, using two complementary measures: a kriging-based measure of school-place density based on straight-line distance, and a walking accessibility measure based on actual network travel time. By comparing the two measures across 16,167 Census Small Areas and modelling their association with active travel in urban areas, the analysis tested the idea that opportunities for walking and cycling to school may depend on where children live. 

The two measures agreed on average but diverged in many individual areas. These divergences consistently reflected severance, where physical barriers such as motorways, rivers, and stretches of water cut an area off from schools that are nearby in straight-line terms but not reachable on foot. The Jacob's Island Census Small Area in Cork (047132051) illustrates this clearly: it carries a kriged school-place density of 2,303 per 100 children-above the 95th percentile nationally among urban CSAs and yet its walking accessibility is exactly zero because the Mahon estuary and the South Ring Road make every nearby school unreachable within 30 minutes on foot. This pattern was not isolated: across the urban modelling sample, 82 Census Small Areas had zero walking accessibility despite above-median kriged density, of which Table 1 lists the fifteen most extreme cases, and the bivariate map showed these severance areas concentrated along the coast, on peninsulas, and around estuaries at national scale.

The statistical model reinforced this picture. The walking accessibility measure, based on network travel time along the actual road and path network, had a strong and fairly certain association with active travel: predicted active travel rose from around 15\% at the lowest levels of walking accessibility to around 60\% at the highest, with a narrow credible interval throughout. The kriging measure, based on straight-line distance, had only a small and uncertain association, with predicted active travel varying within a band of roughly 41\% to 44\% across its full range. Both models converged well, with maximum $\hat{R}$ values of 1.002 and 1.001 respectively and no divergent transitions.

Taken together, these results suggest that what matters most for active school travel is not how close schools are in straight-line terms, but whether they can actually be reached on foot. This has practical implications for how school accessibility is measured and for how active travel is supported: network-based measures capture the real opportunity for walking more faithfully than straight-line measures, and local infrastructure that overcomes barriers, pedestrian crossings, footbridges, and safe routes around estuaries and motorways perhaps are as important as the broader distribution of school places in shaping opportunities for active school travel.

This analysis focused on walking accessibility; several natural extensions follow from it. First, the travel-time measure used a step decay function with a hard 30-minute cut-off, treating all reachable school places equally regardless of travel time. Applying a continuous decay function so that more distant school places contribute less to the score, while extending the threshold to 45 minutes, would make the travel-time measure more comparable in 
catchment to the kriging measure, which captures school-place density over a wider area with an implicit decay through the variogram structure; both adjustments are supported within the r5r framework: the decay function through the \texttt{decay\_function} argument and the extended threshold through the \texttt{cutoff} parameter in \texttt{r5r::accessibility} [@pereira2021r5r]. Second, the fifteen Census Small Areas in Celbridge-Leixlip on the south-western Dublin fringe --- all carrying raw kriging values of 498--505 and sitting above the 99th percentile of the national distribution despite normal child populations and only a single nearby school point --- point to over-smoothing from the long-range Gaussian variogram fitted here; reducing the variogram range and increasing the nugget would allow local variation in dense urban-fringe areas to be better resolved. Third, separating the analysis by school level would allow primary and post-primary accessibility to be matched more precisely to the age groups each measure is intended to serve. Finally, repeating the analysis at settlement level would contextualise these Census Small Area findings within the broader town-scale patterns examined in previous Irish research on school provision and active transport [@mancini2025census].

\newpage

# References {-}