---
title: "Distance, Mobility, and Mode Choice: A Spatial Analysis of School Travel Behaviour in Ireland"
author:
  - name: Judith Beltrán López
  - name: Sananthaa Jagadesan Senthamaraikannan
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "90pt"
author_textsize: "60pt"
body_textsize: "25pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 18pt;
}
.leaflet-container {
  width: 100% !important;
  height: 460px !important;
  border-radius: 12px;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(leaflet)
library(htmltools)
library(janitor)
library(sf)
library(scales)
```

# Background and Problem
- Daily travel to school significantly influences physical activity levels in children and adolescents  
- Increased use of unhealthy transport has reduced walking and cycling  
- Active travel provides important opportunities for regular physical activity  
- In Ireland, school location, distance to school, and neighbourhood features affect travel choices  
- Travel habits formed in childhood often continue into adulthood  


# Objectives of Project

1.  Map and analyse the spatial distribution of schools across Ireland.
2.  Examine the distance between students’ residential locations and their schools, and investigate student mobility related to school attendance.
3.  Analyse school travel behaviour and assess the prevalence of healthy versus less healthy commuting modes.
4. Explore the relationship between school proximity, travel mode, and potential health implications.

# Data Sources and Datasets
Data on schools in Ireland are of the following Types:

-School-Level Data (Department of Education): School type, enrolment, and geographic location (Eircode,latitude,       longitude).
-Settlement-Level Travel Data (CSO): Census data on modes of travel to education at settlement level across multiple years.
-Spatial Boundary Data (CSO): Geographic boundaries for towns and urban areas used for spatial mapping and analysis.
<!-- This #### below causes the columns to break -->

<!-- You can change where you put these breaks, and the "column_numbers: 4" argument in the YAMLS header to change the layout of the poster, with care! -->

<!--####-->

# Early Results 

The analysis began with an exploratory examination of the spatial distribution of schools across Ireland using maps and density plots. This helped validate the dataset, reveal spatial patterns, and highlight regional differences in school availability. The results show a clear urban–rural divide, with Dublin emerging as the main concentration of schools, suggesting shorter travel distances and greater potential for active transport, while rural areas are more likely to depend on motorised transport due to longer distances.

```{r standard-plot1, out.width='80%', fig.align='center', fig.cap='Great figure!', fig.height=4,warning=FALSE}
library(janitor)
library(tidyverse)

df_clean <- read_csv("../data/SAP2022T11T1TOWN22.20251128T151123.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  filter(means_of_travel != "Not stated") %>%
  mutate(
    health_group = if_else(means_of_travel %in% c("On Foot", "Bicycle"),
                           "Healthy", "Not healthy")
  )

df_clean %>%
  group_by(means_of_travel) %>%
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(percentage = total / sum(total) * 100) %>%
  ggplot(aes(x = reorder(means_of_travel, percentage), y = percentage)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "Percentage of Usually Resident Population by Travel Mode",
    x = "Means of Travel",
    y = "Percentage (%)"
  )
```

```{r live-map, out.width='100%', fig.height=4.7, message=FALSE, warning=FALSE}
schools_df <- readr::read_csv("../data/School_Data_Eircode_2024_2025.csv",
  locale = readr::locale(encoding = "Latin1"),
  show_col_types = FALSE
) |>
  janitor::clean_names()

schools_sf <- schools_df |>
  filter(!is.na(school_latitude), !is.na(school_longitude)) |>
  st_as_sf(
    coords = c("school_longitude", "school_latitude"),
    crs = 4326,
    remove = FALSE
  )

enrol <- schools_sf$total_2024_2025
enrol[is.na(enrol)] <- 0
radius <- rescale(enrol, to = c(2, 9))

pal <- colorFactor(
  palette = "Dark2",
  domain  = schools_sf$county,
  na.color = "#999999"
)

leaflet(schools_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~school_longitude,
    lat = ~school_latitude,
    radius = radius,
    stroke = FALSE,
    fillOpacity = 0.75,
    color = ~pal(county),
    popup = ~HTML(paste0(
      "<b>", official_school_name, "</b><br>",
      "County: ", county, "<br>",
      ifelse(is.na(total_2024_2025),
             "",
             paste0("Enrolment: ", total_2024_2025))
    ))
  ) |>
  addLegend(
    "bottomright",
    pal = pal,
    values = ~county,
    title = "County"
  ) |>
  fitBounds(
    lng1 = min(schools_sf$school_longitude, na.rm = TRUE),
    lat1 = min(schools_sf$school_latitude,  na.rm = TRUE),
    lng2 = max(schools_sf$school_longitude, na.rm = TRUE),
    lat2 = max(schools_sf$school_latitude,  na.rm = TRUE)
  )
```

Kriging is a geostatistical interpolation technique that uses spatial relationships to estimate values at unsampled locations, helping to reveal spatial patterns and inequalities in school locations, travel behaviour, and related health impacts.


```{r standard-plot3, out.width='90%', fig.align='center', fig.cap='Kriging schools in Co Galway', fig.height=4,warning=FALSE,message=FALSE}

library(readxl)
library(dplyr)
library(sf)
library(sp)
library(gstat)
library(ggplot2)
library(viridis)

schools_df <- read_excel("../data/Data_on_Individual_Schools_Mainstream_2024_25.xlsx", sheet = 2) %>%
  rename(
    Roll_Number = `Roll Number`,
    longitude   = `School Longitude`,
    latitude    = `School Latitude`
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  mutate(
    enrolment = suppressWarnings(as.numeric(`Enrolment per Return`))
  ) %>%
  filter(is.finite(enrolment), enrolment > 0) %>%     # avoid log(0) and NA
  mutate(log_enrol = log(enrolment)) %>%
  filter(is.finite(log_enrol))

ireland_counties <- st_read("../data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE)

galway_sf <- ireland_counties %>%
  filter(NAME_1 == "Galway") %>%
  st_transform(2157)

schools_sf <- st_as_sf(
  schools_df,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
) %>%
  st_transform(2157)

schools_galway_sf <- schools_sf[galway_sf, ] %>%
  filter(is.finite(log_enrol))

schools_sp <- as(schools_galway_sf, "Spatial")



grid_centers <- st_make_grid(galway_sf, cellsize = 5000, what = "centers")
grid_sf <- st_sf(grid_id = seq_along(grid_centers), geometry = grid_centers)
grid_sf <- grid_sf[galway_sf, ]
grid_sp <- as(grid_sf, "Spatial")

stopifnot(length(grid_sp) > 0)


df_sp <- as.data.frame(schools_sp)
xy <- coordinates(schools_sp)
df_sp$x <- xy[, 1]
df_sp$y <- xy[, 2]

dup_n <- sum(duplicated(df_sp[, c("x", "y")]))


schools_agg <- df_sp %>%
  group_by(x, y) %>%
  summarise(log_enrol = mean(log_enrol, na.rm = TRUE), .groups = "drop")

coordinates(schools_agg) <- ~ x + y
proj4string(schools_agg) <- proj4string(schools_sp)

stopifnot(nrow(as.data.frame(schools_agg)) > 10)


v <- variogram(log_enrol ~ 1, data = schools_agg, width = 5000, cutoff = 60000)

v_init <- vgm(psill = 0.8, model = "Exp", range = 30000, nugget = 0.1)
fv <- fit.variogram(v, v_init)

kriged_sp <- invisible(suppressMessages(krige(
  formula   = log_enrol ~ 1,
  locations = schools_agg,
  newdata   = grid_sp,
  model     = fv,
  nmax      = 100,   
  nmin      = 1,     
  maxdist   = 1e6    
)))

kriged_sp$enrol_pred <- exp(kriged_sp$var1.pred)
kriged_sp$sd_log     <- sqrt(kriged_sp$var1.var)



kriged_sf <- st_as_sf(kriged_sp)
schools_sf_plot <- st_as_sf(schools_agg)



coords <- st_coordinates(kriged_sf)

kriged_df <- cbind(
  as.data.frame(kriged_sf),
  x = coords[,1],
  y = coords[,2]
)

ggplot() +
  geom_tile(
    data = kriged_df,
    aes(x = x, y = y, fill = enrol_pred),
    width = 5000,
    height = 5000
  ) +
  geom_sf(data = galway_sf, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = schools_sf_plot, color = "red", size = 0.6) +
  scale_fill_viridis_c(name = "Pred enrolment") +
  coord_sf(crs = st_crs(galway_sf)) +
  labs(title = "Kriged Enrolment Surface – County Galway") +
  theme_bw()

```
```{r standard-plot4, out.width='90%', fig.align='center', fig.cap='Kriging schools in Co Galway', fig.height=4,warning=FALSE,message=FALSE}

ggplot() +
  geom_tile(
    data = kriged_df,
    aes(x = x, y = y, fill = var1.var),
    width = 5000,
    height = 5000
  ) +
  geom_sf(data = galway_sf, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = schools_sf_plot, color = "red", size = 0.6) +
  scale_fill_viridis_c(name = "Kriging variance", trans = "log") +
  coord_sf(crs = st_crs(galway_sf)) +
  labs(title = "Kriging Uncertainty – County Galway") +
  theme_bw()

```
#### 

# Next Project Steps
-Define outcome variables (e.g. enrolment size, accessibility, resource availability)
-Control for confounding factors (socioeconomic context, location, school size, demographics)
-Apply county-level kriging to examine spatial variation in school enrolment
-Extend analysis using regression or multivariable models to assess relationships between distance, travel mode, and    school characteristics

# GitHub
This project can be viewed at our GitHub repository here: <https://github.com/judithbeltranlopez/Final-Project>

# References



