---
title: "Distance, Mobility, and Mode Choice: A Spatial Analysis of School Travel Behaviour in Ireland"
author:
  - name: Judith Beltrán López
  - name: Sananthaa Jagadesan Senthamaraikannan
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "70pt"
author_textsize: "45pt"
body_textsize: "25pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 16pt;
}
.leaflet-container {
  width: 100% !important;
  height: 460px !important;
  border-radius: 12px;
}

.bib {
  font-size: 18pt;
  line-height: 1.15;
}

.bib a {
  word-break: break-all;
  overflow-wrap: anywhere;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(leaflet)
library(htmltools)
library(janitor)
library(sf)
library(sp)
library(gstat)
library(ggplot2)
library(viridis)
library(readxl)
library(dplyr)
library(scales)
library(readr)
```

# Background 
- Daily travel to school significantly influences physical activity levels in children and adolescents  
- Increased use of unhealthy transport has reduced walking and cycling  
- Active travel provides important opportunities for regular physical activity  
- In Ireland, school location, distance to school, and neighbourhood features affect travel choices  
- Travel habits formed in childhood often continue into adulthood  


# Objectives 

1.  Map and analyse the spatial distribution of schools across Ireland.
2.  Examine the distance between students’ residential locations and their schools, and investigate student mobility related to school attendance.
3.  Analyse school travel behaviour and assess the prevalence of healthy versus less healthy commuting modes.
4. Explore the relationship between school proximity, travel mode, and potential health implications.

# Datasets
Data on schools in Ireland are of the following Types:

- School-Level Data (Department of Education): School type, enrolment, and geographic location (Eircode,latitude,       longitude).
- Settlement-Level Travel Data (CSO): Census data on modes of travel to education at settlement level across multiple years.
- Spatial Boundary Data (CSO): Geographic boundaries for towns and urban areas used for spatial mapping and analysis.
<!-- This #### below causes the columns to break -->

<!-- You can change where you put these breaks, and the "column_numbers: 4" argument in the YAMLS header to change the layout of the poster, with care! -->

<!--####-->

# Early Results 

```{r live-map, out.width='95%', fig.height=5, message=FALSE, warning=FALSE, fig.cap='Spatial distribution of schools across Ireland'}
schools_df <- readr::read_csv("../data/School_Data_Eircode_2024_2025.csv",
  locale = readr::locale(encoding = "Latin1"),
  show_col_types = FALSE
) |>
  janitor::clean_names()

schools_sf <- schools_df |>
  dplyr::filter(!is.na(school_latitude), !is.na(school_longitude)) |>
  sf::st_as_sf(coords = c("school_longitude", "school_latitude"), crs = 4326, remove = FALSE)

enrol <- schools_sf$total_2024_2025
enrol[is.na(enrol)] <- 0
radius <- scales::rescale(enrol, to = c(2, 9))

ireland_counties <- sf::st_read("../data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE) |>
  sf::st_transform(4326)

county_counts <- schools_df |>
  janitor::clean_names() |>
  dplyr::filter(!is.na(county)) |>
  dplyr::count(county, name = "n_schools")

ireland_counties <- ireland_counties |>
  dplyr::left_join(county_counts, by = c("NAME_1" = "county")) |>
  dplyr::mutate(n_schools = replace_na(n_schools, 0),
                hover_label = paste0(NAME_1, ": ", n_schools, " schools"))

leaflet(
  options = leafletOptions(zoomControl = TRUE, scrollWheelZoom = TRUE)
) |>
  addProviderTiles(providers$CartoDB.Positron) |>

  addPolygons(
    data = ireland_counties,
    fillOpacity = 0,
    color = "black",
    weight = 1,
    opacity = 0.8,
    label = ~hover_label,
    labelOptions = labelOptions(
      direction = "auto",
      textsize = "14px",
      style = list("font-weight" = "bold", "padding" = "6px 8px")
    ),
    highlightOptions = highlightOptions(
      weight = 3,
      color = "black",
      bringToFront = TRUE
    )
  ) |>

  addCircleMarkers(
    data = schools_sf,
    lng = ~school_longitude,
    lat = ~school_latitude,
    radius = radius,
    stroke = FALSE,
    fillOpacity = 0.65,
    color = "#a80050",
    popup = NULL
  ) |>

  fitBounds(
    lng1 = min(schools_sf$school_longitude, na.rm = TRUE),
    lat1 = min(schools_sf$school_latitude,  na.rm = TRUE),
    lng2 = max(schools_sf$school_longitude, na.rm = TRUE),
    lat2 = max(schools_sf$school_latitude,  na.rm = TRUE)
  )

```
Kriging is a geostatistical interpolation technique that uses spatial relationships to estimate values at unsampled locations, helping to reveal spatial patterns and inequalities in school locations, travel behaviour, and related health impacts.
```{r standard-plotireland, out.width='95%', fig.align='center', fig.cap='Kriging schools Ireland', fig.height=4,warning=FALSE,message=FALSE, results="hide"}

cell <- 5000  

schools_df <- read_excel("../data/Data_on_Individual_Schools_Mainstream_2024_25.xlsx", sheet = 2) %>%
  rename(
    longitude = `School Longitude`,
    latitude  = `School Latitude`
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  mutate(enrolment = suppressWarnings(as.numeric(`Enrolment per Return`))) %>%
  filter(is.finite(enrolment), enrolment > 0) %>%
  mutate(log_enrol = log(enrolment)) %>%
  filter(is.finite(log_enrol))

ireland_counties <- st_read("../data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE)

ireland_sf <- st_union(st_transform(ireland_counties, 2157)) %>%
  st_as_sf() %>%
  mutate(id = 1)

schools_sf <- st_as_sf(schools_df, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(2157)

schools_irl_sf <- schools_sf[ireland_sf, ] %>%
  filter(is.finite(log_enrol))

schools_sp <- as(schools_irl_sf, "Spatial")

df_sp <- as.data.frame(schools_sp)
xy <- coordinates(schools_sp)
df_sp$x <- xy[, 1]
df_sp$y <- xy[, 2]

schools_agg <- df_sp %>%
  group_by(x, y) %>%
  summarise(log_enrol = mean(log_enrol, na.rm = TRUE), .groups = "drop")

coordinates(schools_agg) <- ~ x + y
proj4string(schools_agg) <- proj4string(schools_sp)

stopifnot(nrow(as.data.frame(schools_agg)) > 50)

grid_polys <- st_make_grid(ireland_sf, cellsize = cell, what = "polygons")
grid_polys <- st_sf(id = seq_along(grid_polys), geometry = grid_polys)

grid_polys <- st_filter(grid_polys, ireland_sf, .predicate = st_intersects)
stopifnot(nrow(grid_polys) > 0)

grid_centroids_sf <- st_centroid(grid_polys)
grid_centroids_sp <- as(grid_centroids_sf, "Spatial")

v <- variogram(log_enrol ~ 1, data = schools_agg, width = cell, cutoff = 200000)  # 200 km

fv <- fit.variogram(
  v,
  vgm(psill = 0.8, model = "Exp", range = 80000, nugget = 0.1)  # start range ~80 km
)

kriged_sp <- krige(
  log_enrol ~ 1,
  locations = schools_agg,
  newdata   = grid_centroids_sp,
  model     = fv,
  nmax      = 150,
  nmin      = 1,
  maxdist   = 1e6
)

kriged_sp$enrol_pred <- exp(kriged_sp$var1.pred)

kriged_centroids_sf <- st_as_sf(kriged_sp)

grid_polys$enrol_pred <- kriged_centroids_sf$enrol_pred
grid_polys$krig_var   <- kriged_centroids_sf$var1.var

schools_sf_plot <- st_as_sf(schools_agg)

ggplot() +
  geom_sf(data = grid_polys, aes(fill = enrol_pred), color = NA) +
  geom_sf(data = ireland_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_viridis_c(name = "Pred enrolment") +
  coord_sf(crs = st_crs(ireland_sf)) +
  labs(title = "Kriged Enrolment Surface – Ireland") +
  theme_bw()+
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

```

```{r standard-plot3, out.width='95%', fig.align='center', fig.cap='Kriging schools in Co Galway', fig.height=4,warning=FALSE,message=FALSE, results="hide"}

cell <- 5000 
schools_df <- read_excel("../data/Data_on_Individual_Schools_Mainstream_2024_25.xlsx", sheet = 2) %>%
  rename(
    longitude = `School Longitude`,
    latitude  = `School Latitude`
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  mutate(enrolment = suppressWarnings(as.numeric(`Enrolment per Return`))) %>%
  filter(is.finite(enrolment), enrolment > 0) %>%
  mutate(log_enrol = log(enrolment)) %>%
  filter(is.finite(log_enrol))


ireland_counties <- st_read("../data/gadm41_IRL_shp/gadm41_IRL_2.shp", quiet = TRUE)

galway_sf <- ireland_counties %>%
  filter(NAME_1 == "Galway") %>%
  st_transform(2157)


schools_sf <- st_as_sf(schools_df, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(2157)

schools_galway_sf <- schools_sf[galway_sf, ] %>%
  filter(is.finite(log_enrol))


schools_sp <- as(schools_galway_sf, "Spatial")

df_sp <- as.data.frame(schools_sp)
xy <- coordinates(schools_sp)
df_sp$x <- xy[, 1]
df_sp$y <- xy[, 2]

schools_agg <- df_sp %>%
  group_by(x, y) %>%
  summarise(log_enrol = mean(log_enrol, na.rm = TRUE), .groups = "drop")

coordinates(schools_agg) <- ~ x + y
proj4string(schools_agg) <- proj4string(schools_sp)

stopifnot(nrow(as.data.frame(schools_agg)) > 10)

grid_polys <- st_make_grid(galway_sf, cellsize = cell, what = "polygons")
grid_polys <- st_sf(id = seq_along(grid_polys), geometry = grid_polys)

grid_polys <- st_filter(grid_polys, galway_sf, .predicate = st_intersects)



stopifnot(nrow(grid_polys) > 0)


grid_centroids_sf <- st_centroid(grid_polys)
grid_centroids_sp <- as(grid_centroids_sf, "Spatial")

v <- variogram(log_enrol ~ 1, data = schools_agg, width = cell, cutoff = 60000)

fv <- fit.variogram(
  v,
  vgm(psill = 0.8, model = "Exp", range = 30000, nugget = 0.1)
)

kriged_sp <- krige(
  log_enrol ~ 1,
  locations = schools_agg,
  newdata   = grid_centroids_sp,
  model     = fv,
  nmax      = 100,
  nmin      = 1,
  maxdist   = 1e6
)

kriged_sp$enrol_pred <- exp(kriged_sp$var1.pred)

kriged_centroids_sf <- st_as_sf(kriged_sp)



grid_polys$enrol_pred <- kriged_centroids_sf$enrol_pred
grid_polys$krig_var   <- kriged_centroids_sf$var1.var

schools_sf_plot <- st_as_sf(schools_agg)

ggplot() +
  geom_sf(data = grid_polys, aes(fill = enrol_pred), color = NA) +
  geom_sf(data = galway_sf, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = schools_sf_plot, color = "red", size = 0.6) +
  scale_fill_viridis_c(name = "Pred enrolment") +
  coord_sf(crs = st_crs(galway_sf)) +
  labs(title = "Kriged Enrolment Surface – County Galway") +
  theme_bw()+
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

```


```{r standard-plot5, out.width='95%', fig.align='center', fig.cap='Histogram of active transport in Ireland', fig.height=4,warning=FALSE,message=FALSE}
urb_rural_df <- read_csv("../data/SAPS_2022_Small_Area_UR_171024.csv")%>%mutate(
    category = str_extract(UR_Category_Desc, "(?<=\\.).*"))%>%select("GEOGID","category", "T11_1_FSCCC","T11_1_BISCCC","T11_1_TSCCC","T11_1_NSSCCC","T11_1_WMFHSCCC")%>%distinct()%>%  mutate(
    urban_rural = case_when(
      category %in% c(
        " Cities",
        " Satellite Urban Towns",
        " Independent urban towns"
      ) ~ "Urban",
      
      category %in% c(
        " Rural areas with high urban influence",
        " Rural areas with moderate urban influence",
        " Highly rural/remote areas"
      ) ~ "Rural",
      
      TRUE ~ NA_character_
    )
  ) %>%
  rename(geogid = GEOGID)%>%
  mutate(
    denom = T11_1_TSCCC - T11_1_WMFHSCCC - T11_1_NSSCCC,
    perc_active = ifelse(
      denom > 0,
      100 * (T11_1_FSCCC + T11_1_BISCCC) / denom,
      NA_real_
    )
  )

binwidth <-  5

ggplot(urb_rural_df, aes(x = perc_active)) +
  geom_histogram(
    aes(y = after_stat(count)),
    binwidth = binwidth,
    boundary = 0,
    closed = "left",
    alpha = 0.6
  ) +
  geom_density(
    aes(y = after_stat(density * nrow(urb_rural_df) * 5)),
    linewidth = 1
  ) +
  labs(
    x = "Percentage using active transport (walking + cycling)",
    y = "Number of areas",
    title = "Distribution of Active Transport to School/College/Childcare"
  ) +
  theme_minimal()

```
```{r standard-plot6, out.width='95%', fig.align='center', fig.cap='Histogram of active transport in Ireland splitted in Rural and Urban places', fig.height=4,fig.asp=0.9, warning=FALSE,message=FALSE}

plot_df <- urb_rural_df %>%
  filter(
    !is.na(perc_active),
    perc_active >= 0,
    perc_active <= 100,
    !is.na(urban_rural)
  )

ggplot(plot_df, aes(x = perc_active)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 5,
    boundary = 0,
    closed = "left",
    alpha = 0.5
  ) +
  geom_density(
    aes(color = category),
    trim = TRUE,
    linewidth = 1,
    na.rm = TRUE
  ) +
  facet_wrap(~ urban_rural, nrow = 2) +
  labs(
    x = "Percentage using active transport (walking + cycling)",
    y = "Density",
    title = "Active Transport Distribution by Urban/Rural",
    color = "Category"
  ) +
  theme_minimal()

```
#### 

# Next Project Steps
- Define predictor variables (e.g. enrolment size, accessibility, resource availability)
- Control for confounding factors (socioeconomic context, location, school size, demographics)
- Apply county-level kriging to examine spatial variation in school enrolment
- Extend analysis using regression or multivariable models to assess relationships between distance, travel mode, and    school characteristics

# GitHub
This project can be viewed at our GitHub repository here: <https://github.com/judithbeltranlopez/Final-Project>

# Bibliography

<div class="bib">

Data:

- Department of Education (Ireland). *Data on Individual Schools*.  
  https://www.gov.ie/en/department-of-education/collections/data-on-individual-schools/
- Central Statistics Office (CSO). *SAP2022T11T1TOWN22*.  
  https://data.cso.ie/table/SAP2022T11T1TOWN22
- Central Statistics Office (CSO). *SAP2016T11T1SET*.  
  https://data.cso.ie/table/SAP2016T11T1SET
- Central Statistics Office (CSO). *E6013*.  
  https://data.cso.ie/table/E6013
- Central Statistics Office (CSO). *Census 2022 Small Area Population Statistics*.  
  https://www.cso.ie/en/census/census2022/census2022smallareapopulationstatistics/

Research Context:

- Why are teens half as likely to walk or cycle to school in Dublin as Dunshaughlin?
  https://www.rte.ie/brainstorm/2024/0209/1415046-teens-schools-ireland-active-transport-cycling-walking-educational-provision/
- Children from these communities’: unequal school provision, segregation, and the Irish educational landscape
  https://www.tandfonline.com/doi/full/10.1080/03323315.2022.2118152

